---
title: "Simulation Activity: Principal Component Analysis"
subtitle: "Math 437"
format: html
editor: source
---

```{r}
#| label: load packages
#| message: false
#| warning: false

library(LearnPCA)
library(Lahman)
library(tidymodels)
```


## The Revenge of Math 307

Facts and theorems that either we encountered in Math 307 or depend on terms we learned in Math 307:

- A real square matrix A is orthogonal if $A^T A = A A^T = I$.
- Two matrices A and B are *orthogonally equivalent* if there exists an orthogonal matrix $P$ such that $A = P^T B P$.
- A real $p \times p$ matrix A is symmetric iff A is orthogonally equivalent to a real diagonal matrix.
- Covariance and correlation matrices are real symmetric $p \times p$ matrices.

The result of these facts and theorems is that *any* covariance or correlation matrix can be written in the form $P^T D P$, where $P$ is an orthogonal $p \times p$ matrix and $D$ is a real diagonal matrix.

The following theorems are presented without proof, but critical for understanding Principal Component Analysis.

1. The diagonal entries of $D$ are the eigenvalues of the covariance matrix $\Sigma$ and are non-negative.
2. The columns of $P$ are the eigenvectors of $\Sigma$.
3. The transformation $X' = P^T X$ transforms the standard ordered basis $\{x_1, x_2, \ldots, x_p\}$ to a new ordered basis $\{x_1', x_2', \ldots, x_p'\}$.


```{r}
#|label: simulated data 

set.seed(109)
a1 <- rnorm(1000)
a2 <- rnorm(1000)
simdata <- data.frame(
  x1 = 2*a1 - a2,
  x2 = a1 + a2
)

plot(x2 ~ x1, data = simdata)

```

```{r}
#| label: covariance matrix

cov_matrix <- cov(simdata)
cov_matrix

```

```{r}
#| label: eigenvalues of covariance matrix

eigen(cov_matrix)

```

Notice that the eigenvectors form an *ordered* basis where the order matches the eigenvalues in *descending* order.

Is P orthogonal?

```{r}
#| label: orthogonality

D <- diag(eigen(cov_matrix)$values)
P <- eigen(cov_matrix)$vectors

apply(P, 2, function(x) sqrt(sum(x^2)))     # check that columns have norm 1

solve(P)     # P^-1
t(P)     # P^T

```

The columns of P have norm 1 and (at least to the precision that we are displaying) $P^T$ and $P^{-1}$ are equivalent. So $P$ is orthogonal.

Note that the `eigen()` function defines the `$vectors` matrix $P$ such that $D = P^T A P$, and so to reconstruct the covariance matrix:

```{r}
#| label: reconstruct covariance matrix
# reconstruct using P D P^T
P %*% D %*% t(P)
cov_matrix

```

We can thus create the vector $(x_1', x_2')$ such that 

$$
x_1' = P[1,1] x_1 + P[1,2] x_2
$$

and

$$
x_2' = P[2,1] x_1 + P[2,2] x_2
$$

This gets a bit tricky in the code because $(x_1, x_2)$ is represented as a row vector:

```{r}
#| label: convert x to x'

simdata2 <- as.matrix(simdata) %*% P
plot(simdata2[,1], simdata2[,2], xlab = "x1'", ylab = "x2'")

```

```{r}
#| label: check covariance now

cov(simdata2)

```

1. What do you notice about this graph of `simdata2` compared to the original graph of `simdata`?

### Why Does This Matter?

```{r}
#| label: total variation

x1_var <- var(simdata$x1)
x2_var <- var(simdata$x2)
total_var1 <- x1_var + x2_var
# really should be sum of squares but the 1/(n-1) is just factored out

x1_prime_var <- var(simdata2[,1])
x2_prime_var <- var(simdata2[,2])
total_var2 <- x1_prime_var + x2_prime_var

c(total_var1, total_var2)

```

Notice that the total amount of variance in the data is the same whether we choose (x1, x2) or (x1', x2') as our basis. But:

```{r}
#| label: proportion of variance

c(x1_var, x2_var)/total_var1
c(x1_prime_var, x2_prime_var)/total_var2

```

By rotating our basis to (x1', x2'), we can maximize the proportion of variance attributed to variation along the direction defined by the first vector in our ordered basis. 

Moreover, we can exactly reconstruct these maximized proportions by normalizing the *eigenvalues* so that they sum to 1:

```{r}
#| label: normalize the eigenvalues

D/sum(D)     # diagonal matrix with % variation on diagonal

```

The `LearnPCA` package has a nice visualization of this rotation:

```{r}
#| label: LearnPCA visual of the basis rotation

# comment this out if it doesn't work or if you want to render later
LearnPCA::PCsearch()
```

Note that each of the vectors in the new ordered basis is a linear combination of $x_1, x_2, \ldots, x_p$. So PCA does **NOT** perform variable selection. What it *does* do is change the basis to more effectively partition the "signal" vs. "noise" when dealing with  highly (positively or negatively) correlated variables. 

## Dimensionality Reduction

The advantage of this partition is that lower-indexed eigenvectors in the basis represent interpretable "signals" and higher-indexed eigenvectors represent uninterpretable "noise." Therefore, we can get rid of the eigenvectors that we believe entirely describe the "noise", reducing the dimensionality of our predictor space.

For example, suppose that we wanted to project our simulated (x1, x2) data into one-dimensional space. If we tried to do this with just x1:

```{r}
#| label: attempt to reconstruct from x1 and x2
z <- simdata[,1] # only the first column

par(mfcol = c(1,2))
plot(simdata$x1, z, xlab = "real x1", ylab = "reconstituted x1",
     main = paste0("Correlation: ", round(cor(simdata$x1, z), 3)))
abline(a = 0, b = 1, col = "red")
plot(simdata$x2, z, xlab = "real x2a", ylab = "reconstituted x2a",
     main = paste0("Correlation: ", round(cor(simdata$x2, z), 3)))
abline(a = 0, b = 1, col = "red")
```

We have preserved all information about `x1` but lost a ton of information about `x2`. On the other hand, if we use the direction specified by the first eigenvector in our ordered basis:

```{r}
#| label: reconstruct based on x1 prime

z <- simdata2[,1] # only the first column
P1 <- P[,1] # only the first column of the transformation

reconstituted <- z %*% t(P1)

par(mfcol = c(1,2))
plot(simdata$x1, reconstituted[,1], xlab = "real x1", ylab = "reconstituted x1",
     main = paste0("Correlation: ", round(cor(simdata$x1, reconstituted[,1]), 3)))
abline(a = 0, b = 1, col = "red")
plot(simdata$x2, reconstituted[,2], xlab = "real x2", ylab = "reconstituted x2",
     main = paste0("Correlation: ", round(cor(simdata$x2, reconstituted[,2]), 3)))
abline(a = 0, b = 1, col = "red")
```

We can see that we are still doing a very good job of preserving the information contained in `x1`, while doing a much-less-awful job of preserving the information contained in `x2`.
