---
title: "Activity: Hierarchical and Model-Based Clustering"
subtitle: "Math 437"
format: html
editor: source
---

## New Packages

```{r}
#| label: load package
#| message: false
#| warning: false
library(tidyverse)
library(tidymodels)
library(tidyclust)
library(dendextend)
library(mclust)
library(factoextra)
```

Note the new packages: `dendextend` is an awesome package for making the results of your hierarchical clustering analysis "pop", `mclust` does model-based clustering, and `factoextra` makes `ggplot2`-based visualizations for PCA and model-based clustering. 

## Data

Hierarchical clustering is *much* easier to visualize with a relatively small number of observations. So here we'll take a random sample of 60 fruits and vegetables to cluster.

```{r}
#| label: load data
Restrain <- readr::read_csv(here::here("Data/Restrain.csv"))

fv <- Restrain |>
  filter(SubCategory %in% c("Fruits", "Vegetables"))

fv$SubCategory <- factor(fv$SubCategory, levels = c("Fruits","Vegetables"))

set.seed(1880)
fv_split <- initial_split(fv, prop = 0.50) 

fv_train <- training(fv_split)
fv_test <- testing(fv_split)

```

## Hierarchical Clustering Analysis

```{r}
#| label: set up the model

fv_hc_complete <- hier_clust(
  linkage_method = "complete"     # default
)     # model

fv_recipe <- recipe(
  ~ Taste + Healthiness + Cravings, data = fv_train
) |>
  step_normalize(all_numeric_predictors())     # all distances are calculated using z-scores, to even out values since measured in different metrics

fv_hc1 <- workflow() |>
  add_model(fv_hc_complete) |>
  add_recipe(fv_recipe)

```

### Defining the Linkage Method

While the distance between two points is (relatively) straightforward to measure, the distance between two groups of points is not. The `linkage_method` argument indicates how the dissimilarity (generalized "distance") between two clusters is computed.

There are several possible methods that can be used to define the linkage. The most popular of these are:

#### Complete linkage

This is the default in `hier_clust`.

$$
D(k, k') = \max_{x_i \in k, x_i' \in k'}||x_i - x_i'||
$$

Essentially, find all possible pairwise Euclidean distances between the points in the clusters $k$ and $k'$, then find the maximum of those distances.

#### Single linkage

$$
D(k, k') = \min_{x_i \in k, x_i' \in k'}||x_i - x_i'||
$$

Essentially, find all possible pairwise Euclidean distances between the points in the clusters $k$ and $k'$, then find the *minimum* of those distances instead of the maximum.

#### Ward's Method

$$
D(k, k') = ESS_{k \cup k'} - (ESS_{k} + ESS_{k'})
$$

where ESS is the "error sum of squares" (essentially RSS or within-group sum of squares) and $k \cup k'$ is the merged cluster containing both the observations in cluster $k$ and the observations in cluster $k'$. Ward's method has very nice connections with LDA, PCA, and k-means clustering.

Just to be confusing, `hier_clust` will accept either `ward.D` or `ward.D2` as a `linkage_method`. My understanding is that you want to use `ward.D2`. 

#### Other Methods

The textbook also talks about `average` and `centroid` methods. Don't use `centroid` unless you really know what you're doing.

### What About `num_clusters`?

In k-means clustering, there is no guarantee that the best `k`-cluster model has anything to do with the best `(k+1)`-cluster model. However, in hierarchical clustering, as the number of clusters increases, old clusters just split into new ones. That is, if you use `num_clusters = 3` and then decide you want to use `num_clusters = 4`, the only thing that will happen is one of the original 3 clusters will split in 2. Literally the only reason to set a value for `num_clusters` in `hier_clust` is to tell R (after the fact) how many clusters you see in the visualization of the results - typically we keep this at the default `NULL` until we see the results and decide how many clusters to use.

## Visualizing the Results

```{r}
#| label: fit the model

hc_complete_fit <- fit(
  fv_hc1,
  data = fv_train
)

```

A *dendrogram* is a tree-like visualization of the clusters.

```{r}
#| label: plot dendrogram

fv_dendrogram <- hc_complete_fit |>
  extract_fit_engine() |>
  as.dendrogram()

plot(fv_dendrogram)

```


This is annoying to read. Let's make some customizations to the original plot:

```{r}
#| label: customize plot dendrogram

fv_Foods <- fv_train$Food[
  str_remove(labels(fv_dendrogram), "v") |> as.numeric()
]
fv_dendrogram_3clusters <- fv_dendrogram |>
  set("labels", fv_Foods) |>
  set("labels_cex", 0.8) |>
  set("branches_lwd", 0.7) |>
  color_branches(k = 3, col = c("black","blue", "red")) |>
  color_labels(k = 3, col = c("black", "blue", "red"))

plot(fv_dendrogram_3clusters)

```

1. Is Pear (very right of the black cluster) considered more similar to a Tomato (very left) or a Carrot (very left of the blue cluster)? Why?

Tomato, similarity is measured by how soon a cluster is made, y-axis, not by x-axis.

2. Apple (fourth from the left) is sandwiched in between Raspberries (third from left) and Blueberries (fifth from left). Which food is it considered more similar to? Why?

Blueberries as it is immediately clustered with it.

3. With a four-cluster partition, which cluster (black, blue, or red) would get split in two? Why?

Black, it gets cut off from blue and red

Once we determine the number of clusters, we can obtain the centroids (geometric means) of each predictor in each cluster:

```{r}
#| label: extract centroids

hc_complete_centroids <- hc_complete_fit |>
  extract_centroids(num_clusters = 3)

hc_complete_centroids

# Means of each cluster technically?

```

Here all the healthiness scores on the original scale are high and the cravings scores are low, so it's a bit hard to interpret what the clusters have in common. One nice thing about converting to z-scores is that we can still interpret positive vs. negative z-scores the way we usually do as being "relative to average". 

4. Using the general adjectives "high", "average", and "low," explain what the centroids suggest about each cluster.

Note that the cluster assignments don't necessarily match the order on the dendrogram:

```{r}
#| label: extract clusters

cluster_assignments_complete <- hc_complete_fit |>
  extract_cluster_assignment(num_clusters = 3)

fv_train |>
  dplyr::select(Food) |>
  bind_cols(cluster_assignments_complete) |>
  slice(1:10)     # first 10 rows

```

The cluster assignments should be unique up to a rotation, so you may have different cluster-to-color mappings, but generally everything in the black cluster should have the same `Cluster_x` everything in the red cluster should be in the same `Cluster_y`, and everything in the blue cluster should be in the same `Cluster_z`.

The plot below plots the Taste and Healthiness of the foods being clustered, as well as the cluster assignments. The large asterisks show the centroids of the 3 clusters.

```{r}
#| label: plot cluster assignments

undo_normalize <- function(x, orig){
  return(x * sd(orig) + mean(orig))     # undo normalization
}

hc_original_scale <- hc_complete_centroids |>
  mutate(
    Taste = undo_normalize(Taste, orig = fv_train$Taste),
    Healthiness = undo_normalize(Healthiness, orig = fv_train$Healthiness),
    Cravings = undo_normalize(Cravings, orig = fv_train$Cravings)
  )

ggplot(data = fv_train |> bind_cols(cluster_assignments_complete),
       aes(x = Healthiness, y = Taste, color = .cluster)) +
  geom_point() +
  geom_point(data = hc_original_scale, shape = 8, size = 5) +
  scale_color_manual(labels = c("Cluster_1", "Cluster_2", "Cluster_3"),
                     values = c("blue", "black", "red")) +
  theme(legend.position = "none")

# We just denormalized the centroids and plotted onto the same scale plot

```

The tidyclust authors suggest *not* using `predict` on the training set, since it will predict the observations as if they were *new* data (and thus could get predicted to be in a totally different cluster!). However, there is no such issue making predictions on the test set:

```{r}
#| label: make cluster predictions

fv_cluster_preds <- hc_complete_fit |>
  predict(
    new_data = fv_test,
    num_clusters = 3
  )

fv_test |>
  dplyr::select(Food) |>
  bind_cols(cluster_assignments_complete) |>
  slice(1:10)     # first 10 rows

```

Let's see what a three-cluster dendrogram looks like using single linkage:

```{r}
#| label: single linkage instead

```

5. What differences do you notice in the shape of the tree between complete and single linkage?

Let's look at Ward's method:

```{r}
#| label: with Ward method

```

6. Does the dendrogram produced by Ward's method look more like the complete-linkage or single-linkage tree?


## Model-Based Clustering

Model-based clustering is sometimes referred to as "soft k-means clustering." Instead of assigning each observation to a specific cluster, model-based clustering assumes a specific distribution shape within each cluster, automatically finds the optimal number of clusters based on that shape, and finds the predicted probability of each point belonging to each cluster.

This approach makes it much easier to handle outliers. 

### Model-Based Clustering with `mclust`

Model-based clustering with `mclust` assumes a multivariate normal distribution for the set of predictors. Therefore, it is important to convert categorical variables to indicator variables and remove zero-variance variables; however, scaling is unnecessary. Also, the model assumes multivariate normality *within each cluster*, so transformation of the predictors is generally unnecessary as points far out in the tail(s) will get mapped to their own cluster.

Since we only have numerical predictors here, there is no need to pre-process the data.

The `mclust` package searches over many possible combinations of similarities/differences in the shape, size (volume), and orientation of the groups' distributions. Each combination is represented by a three-letter code. See <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5096736/figure/F2/> for an illustration of the different combinations in 2-dimensional space.

### Parameters to Tune

None! Technically there are two parameters to tune:

* The number of clusters
* The shape of the density function within the clusters

However, the `mclust` package automatically searches over the entire parameter space, so we don't need to actually *tune* anything.

### Running the Model

As far as I can tell, the function `gm_clust()` is coming to `tidyclust` eventually to do Gaussian Mixture Model-based clustering in the tidymodels framework, but it is not available in Version 0.2.4.

Luckily there is not much pre-processing to do, and we can use the same tricks we saw in the Chapter 6 lab.

```{r}
#| label: prep with model.matrix

fv_train_predictors <- model.matrix(
  ~ Taste + Healthiness + Cravings, data = fv_train
)[,-1]     # first column represents an intercept

```

```{r}
#| label: cluster with mclust

mclust1 <- Mclust(
  fv_train_predictors,
  G = 1:9     # search over all possible combinations of 1-9 clusters
       # if we want a specific amount just do that
)

summary(mclust1)
plot(mclust1, "BIC")
plot(mclust1, "classification")

```

1. Briefly explain how to read this plot. How was the best model determined?

```{r}
#| label: use tidy to summarize mclust output

tidy(mclust1)

```

2. How do we read this table? What do the foods in each cluster have in common?

```{r}
#| label: classification plot

plot(mclust1, "classification")

```

3. Briefly explain how to read this plot. Based on your answer to Question #2, which cluster is the red cluster? Blue cluster? Green cluster?

4. Notice that there is more "overlap" between the clusters - for example, some blue dots are closer to the red mean than the blue mean by Euclidean distance. Why is this?

```{r}
#| label: add clusters to training set

augmented_mclust1 <- augment(
  mclust1,
  fv_train
)

augmented_mclust1

```

5. What do the `.class` and `.uncertainty` columns indicate?

```{r}
#| label: make uncertainty plot

plot(mclust1, "uncertainty")

```

6. What do bigger and darker circles indicate on this plot?

You can also use the `factoextra` package to make ggplot2-based versions of the same plot:

```{r}
#| label: use fviz_mclust for more plots

fviz_mclust(mclust1, what = "BIC")
fviz_mclust(mclust1, what = "classification")
fviz_mclust(mclust1, what = "uncertainty", palette = "simpsons")     # "rickandmorty" also work

```

The main difference is that the `factoextra` versions of the classification and uncertainty plots use the first two principal components as the x-axis and y-axis instead of creating a `pairs` plot to show the clusters' relationships with each variable.

### Making Predictions with `mclust`

```{r}
#| label: predictions with mclust

```

1. What do the `$classification` and `$z` objects represent?

Unfortunately, there is no `tidy` method available, so we have to augment the test set ourselves.

```{r}
#| label: manually augment with mutate

```

There is *a lot* more that you can do with this package, but this hopefully gives you the basic idea.