---
title: "Activity: k-Nearest Neighbors"
subtitle: "Math 437"
format: html
editor: source
---

```{r}
#| label: load package and data
#| message: false
#| warning: false
library(tidyverse)
library(tidymodels)
library(kknn) # for the actual model fitting
library(forcats) # tidyverse for factors

Restrain <- readr::read_csv(here::here("Data/Restrain.csv"))

Restrain2 <- Restrain |> mutate(SubCategory =
  fct_collapse(SubCategory,
    savory = c("Bakery (savoury)", "Savoury snacks"),
    sweet = c("Bakery (sweet)", "Biscuits", "Confectionery", "Desserts"),
    takeout = c("Takeaway (chain)", "Takeaway (generic)"),
    fruit_veggie = c("Fruits", "Vegetables")
  )
)

set.seed(1880)
Restrain_split <- initial_split(Restrain2, prop = 0.80) 
# can include a strata = "variable_name" argument to do stratified random training/test sampling

Restrain_train2 <- training(Restrain_split)
Restrain_test2 <- testing(Restrain_split)
```


## k-Nearest Neighbors Algorithm

1. Explain how a k-nearest neighbors classification algorithm works.

The model to run k-nearest neighbors is `nearest_neighbor`. Because k-nearest neighbors works with both regression and classification, we must specify the mode. We must also specify the number of neighbors to choose.

```{r}
#| label: knn 

K <- 10     # 10-nearest neighbors
knn_model <- nearest_neighbor(
  mode = "classification",
  neighbors = K,
  dist_power = 2
)

```

2. Briefly explain how Minkowski distance is calculated and what the `dist_power` argument refers to.

```{r}
#| label: knn recipe

knn_recipe <- recipe(
  SubCategory ~ Taste + Cravings + Healthiness,
  data = Restrain_train2
) |>
  step_normalize(all_numeric_predictors()) #|>
# step_dummy(all_nominal_predictors())

```

3. Why should we center and scale the numerical variables?

4. Why do categorical variables need to be converted to indicator variables for this algorithm?

```{r}
#| label: fit knn

knn_wflow <- workflow() |>
  add_model(knn_model) |>
  add_recipe(knn_recipe)

knn_fit <- knn_wflow |>
  fit(data = Restrain_train2)

knn_fit

```

Notice that we are using the `train.kknn` function in the `kknn` package instead of the `knn` function from the `class` package (as done in the ISLR Chapter 12 lab).

### Making Predictions

As usual, we use the `predict` function to make predictions:

```{r}
#| label: make predictions on test set

knn_predictions <- knn_fit |>
  predict(new_data = Restrain_test2)

knn_predictions

```

```{r}
#| label: probabilistic predictions

knn_predictions <- knn_fit |>
  predict(new_data = Restrain_test2, type = "prob")

knn_predictions

```

1. We used 10-nearest neighbors! Why don't we get nice round percentages with these predictions?

We can also use `augment` to add the predictions to the original data frame:

```{r}
#| label: predict with augment

knn_predictions_df <- knn_fit |>
  augment(new_data = Restrain_test2)

knn_predictions_df |>
  dplyr::select(
    Food,
    SubCategory,
    .pred_class,
    .pred_savory,
    .pred_sweet,
    .pred_takeout,
    everything()
  ) |>
  slice(51:54)

```

2. Explain why the "Filet-o-Fish" (row 52) is predicted to be sweet. 

Highest Probability

We can obtain the confusion matrix and see what categories get confused with each other:

```{r}
#| label: confusion matrix

knn_predictions_df |>
  conf_mat(
    truth = SubCategory,
    estimate = .pred_class
  )

```

## Tuning k-Nearest Neighbors

In k-nearest neighbors, we have two hyperparameters that we can tune: the number of neighbors and the distance metric.

Remember that we tune hyperparameters using cross-validation. Here we'll use repeated 5-fold cross-validation.

```{r}
#| label: k-fold cv setup

set.seed(145)

Restrain_kfold <- vfold_cv(Restrain_train2,
                     v = 5, repeats = 2,
                     strata = SubCategory) 
Restrain_kfold

```

Here we'll tune both the number of nearest neighbors and the distance power.

```{r}
#| label: knn with tuning

knn_model_tune <- nearest_neighbor(
  mode = "classification",
  neighbors = tune(),
  dist_power = tune()
)

knn_params <- extract_parameter_set_dials(knn_model_tune)

knn_params

```

Let's set up our search grid using `expand.grid`:

```{r}
#| label: set up parameter grid

knn_param_grid <- expand.grid(
  neighbors = seq(2, 62, by = 10),
  dist_power = c(1, 2, 5)
)

knn_param_grid

```

```{r}
#| label: workflow to tune

knn_wflow_tune <- workflow() |>
  add_model(knn_model_tune) |>
  add_recipe(knn_recipe)

```

Recall that we search over the entire grid using `tune_grid`:

```{r}
#| label: tune knn model

knn_tune <- knn_wflow_tune |>
  tune_grid(
    resamples = Restrain_kfold,
    grid = knn_param_grid,
    metrics = metric_set(roc_auc, brier_class, mn_log_loss)
  )

```

1. Explain the choice of `k` in terms of the bias-variance tradeoff.

```{r}
#| label: investigate tuning results

autoplot(knn_tune)

```

Since we get one tibble out for each fold, we can use `bind_rows` to make the output look a bit nicer and calculate the mean and SE.

```{r}
#| label: unwieldy add error bars to graph
knn_metric_avg <- knn_tune$.metrics |>
  bind_rows(
    .id = "validation_set"
  ) |>
  group_by(neighbors, dist_power, .metric) |>
  summarize(
    mean = mean(.estimate),
    se = sd(.estimate)/sqrt(length(knn_tune$.metrics)),
    .groups = "drop"
  ) |>
  group_by(.metric) |>
  arrange(mean, .by_group = TRUE)

ggplot(knn_metric_avg,
       mapping = aes(
         x = neighbors,
         y = mean,
         color = factor(dist_power)
       )) +
  geom_point() +
  geom_line() +
  geom_errorbar(
    aes(x = neighbors, ymin = mean - se, ymax = mean + se),
    width = 0.2
  ) +
  facet_wrap(
    vars(.metric),nrow = 3,
    scales = "free_y"
  )
```

2. How does the graph of `brier_class` illustrate the bias-variance tradeoff? What combination of `dist_power` and `neighbors` would you select based on this cross-validation?

```{r}
#| label: fit and predict final model

knn_model_final <- nearest_neighbor(
  mode = "classification",
  neighbors = 22,     # hard coded instead of select best is done here
  dist_power = 2
)

knn_wflow_final <- workflow() |>
  add_model(knn_model_final) |>
  add_recipe(knn_recipe)

knn_fit_tuned <- fit(knn_wflow_final,
                     data = Restrain_train2)

knn_predictions_df_tuned <- knn_fit_tuned |>
  augment(new_data = Restrain_test2)

conf_mat(
  knn_predictions_df_tuned,
  truth = SubCategory,
  estimate = .pred_class
)     # some takeout food was confused for fruit or veggie, lets see what it was

knn_predictions_df_tuned |>
  filter(SubCategory == "takeout", .pred_class == "fruit_veggie")

```

