---
title: "Activity: k-Nearest Neighbors"
subtitle: "Math 437"
format: html
editor: source
---

```{r}
#| label: load package and data
#| message: false
#| warning: false
library(tidyverse)
library(tidymodels)
library(kknn) # for the actual model fitting
library(forcats) # tidyverse for factors

Restrain <- readr::read_csv(here::here("Data/Restrain.csv"))

Restrain2 <- Restrain |> mutate(SubCategory =
  fct_collapse(SubCategory,
    savory = c("Bakery (savoury)", "Savoury snacks"),
    sweet = c("Bakery (sweet)", "Biscuits", "Confectionery", "Desserts"),
    takeout = c("Takeaway (chain)", "Takeaway (generic)"),
    fruit_veggie = c("Fruits", "Vegetables")
  )
)

set.seed(1880)
Restrain_split <- initial_split(Restrain2, prop = 0.80) 
# can include a strata = "variable_name" argument to do stratified random training/test sampling

Restrain_train2 <- training(Restrain_split)
Restrain_test2 <- testing(Restrain_split)
```


## k-Nearest Neighbors Algorithm

1. Explain how a k-nearest neighbors classification algorithm works.

The model to run k-nearest neighbors is `nearest_neighbor`. Because k-nearest neighbors works with both regression and classification, we must specify the mode. We must also specify the number of neighbors to choose.

```{r}
#| label: knn model

```

2. Briefly explain how Minkowski distance is calculated and what the `dist_power` argument refers to.

```{r}
#| label: knn recipe

```

3. Why should we center and scale the numerical variables?

4. Why do categorical variables need to be converted to indicator variables for this algorithm?

```{r}
#| label: fit knn

```

Notice that we are using the `train.kknn` function in the `kknn` package instead of the `knn` function from the `class` package (as done in the ISLR Chapter 12 lab).

### Making Predictions

As usual, we use the `predict` function to make predictions:

```{r}
#| label: make predictions on test set

```

```{r}
#| label: probabilistic predictions

```

1. We used 10-nearest neighbors! Why don't we get nice round percentages with these predictions?

We can also use `augment` to add the predictions to the original data frame:

```{r}
#| label: predict with augment

```

2. Explain why the "Filet-o-Fish" (row 52) is predicted to be sweet. 

We can obtain the confusion matrix and see what categories get confused with each other:

```{r}
#| label: confusion matrix

```

## Tuning k-Nearest Neighbors

In k-nearest neighbors, we have two hyperparameters that we can tune: the number of neighbors and the distance metric.

Remember that we tune hyperparameters using cross-validation. Here we'll use repeated 5-fold cross-validation.

```{r}
#| label: k-fold cv setup
set.seed(145)
fv_kfold <- vfold_cv(Restrain_train2,
                     v = 5, repeats = 2,
                     strata = SubCategory) 
fv_kfold
```

Here we'll tune both the number of nearest neighbors and the distance power.

```{r}
#| label: knn with tuning

```

Let's set up our search grid using `expand.grid`:

```{r}
#| label: set up parameter grid

```

```{r}
#| label: workflow to tune

```

Recall that we search over the entire grid using `tune_grid`:

```{r}
#| label: tune knn model

```

1. Explain the choice of `k` in terms of the bias-variance tradeoff.

```{r}
#| label: investigate tuning results

```

Since we get one tibble out for each fold, we can use `bind_rows` to make the output look a bit nicer and calculate the mean and SE.

```{r}
#| label: unwieldy add error bars to graph
knn_metric_avg <- knn_tune$.metrics |>
  bind_rows(
    .id = "validation_set"
  ) |>
  group_by(neighbors, dist_power, .metric) |>
  summarize(
    mean = mean(.estimate),
    se = sd(.estimate)/sqrt(length(knn_tune$.metrics)),
    .groups = "drop"
  ) |>
  group_by(.metric) |>
  arrange(mean, .by_group = TRUE)

ggplot(knn_metric_avg,
       mapping = aes(
         x = neighbors,
         y = mean,
         color = factor(dist_power)
       )) +
  geom_point() +
  geom_line() +
  geom_errorbar(
    aes(x = neighbors, ymin = mean - se, ymax = mean + se),
    width = 0.2
  ) +
  facet_wrap(
    vars(.metric),nrow = 3,
    scales = "free_y"
  )
```

2. How does the graph of `brier_class` illustrate the bias-variance tradeoff? What combination of `dist_power` and `neighbors` would you select based on this cross-validation?

```{r}
#| label: fit and predict final model

```

