---
title: "Activity: Understanding Slow Learning"
subtitle: "Math 437"
format: html
editor: source
---

## A Simple Example

In this example, we're going to build a boosted tree algorithm. We're not going to do any tuning, so it's going to suck; we're using it as a didactic example.

We'll use the `batting_2016` data because it's much easier to explain slow learning with regression problems than classification problems.

```{r}
#| label: create batting_2016 data
#| warning: false
#| message: false
library(Lahman)
library(tidyverse)
library(tidymodels)
library(rpart)

salaries <- Salaries %>%
  dplyr::select(playerID, yearID, teamID, salary)
peopleInfo <- People %>%
  dplyr::select(playerID, birthYear, birthMonth, nameLast,
         nameFirst, bats)
batting <- battingStats() %>% 
  left_join(salaries, 
            by =c("playerID", "yearID", "teamID")) %>%
  left_join(peopleInfo, by = "playerID") %>%
  mutate(age = yearID - birthYear - 
           1L *(birthMonth >= 10)) %>%
  arrange(playerID, yearID, stint)

batting_2016 <- batting %>% filter(yearID == 2016,
                                   !is.na(salary),
                                   G >= 100, AB >= 200
                                   ) %>%
  mutate(salary = log10(salary), # there's a reason for this, I promise
         lgID = factor(lgID, levels = c("AL", "NL"))) # fix the defunct league issue
```

In fast learning, we fit the model one time and whatever we get out - good or bad - is our final model.

In contrast, in slow learning, we start out with terrible models, and on each iteration, we get slightly less terrible, until we (hopefully) have a model that doesn't suck.

Let's arbitrarily fit the model for 50 iterations. In practice, we'd want to tune the number of iterations - large enough that we're making good predictions on the training set, not so large that we're overfitting.

```{r}
#| label: setup predictions
niterations <- 50
salary_predictions <- vector("list", length = niterations+1)
names(salary_predictions) <- paste0("Step", seq(0, niterations))
rmse_valid <- numeric(niterations + 1)
fhat <- vector("list", length = niterations)
```

## Step 1: Initialize the Model

```{r}
#| label: training-validation-test split
set.seed(115, kind = "Mersenne-Twister")
batting_2016_split <- initial_validation_split(
  batting_2016,
  prop = c(.6,     # training
           .2     # validation
           )
)

# 60% of the data used to fit the model
# 20% of the data used to validate the model on each step
# 20% of the data used as a test set

batting_train <- training(batting_2016_split)
batting_test <- testing(batting_2016_split)
batting_valid <- validation(batting_2016_split)

```

### Step 1a: Fit a Useless Model

The book suggests to use $\hat{f}^0(x) = 0$. We can create a much better model while still maintaining the criterion that the model has to be completely useless for predictions: $\hat{f}^0(x) = \bar{y}$. This is the null model that predicts everyone to have the average salary in the training set. If we can't make better predictions than this null model, we are in serious trouble.

```{r}
#| label: Step 1a - useless model

yhat0 <- mean(batting_train$salary)

```

### Step 1b: Compute the Residuals

We now make predictions on the training set using our useless model and store the residuals.

```{r}
#| label: Step 1b - residuals

resid0 <- batting_train$salary - yhat0

batting_new <- batting_train |>
  mutate(
    current_pred = yhat0,
    current_resid = salary - yhat0
  )

```


```{r}
#| label: Step 1 predictions

batting_valid <- batting_valid |>
  mutate(
    current_pred = yhat0,
    current_resid = salary - yhat0
  )
ggplot(batting_valid, aes(x = salary, y = current_pred)) +
  geom_point() +
  labs(x = "Actual Salary", y= "Predicted Salary (Step 0)") +
  geom_abline(intercept = 0, slope = 1)

salary_predictions[[1]] <- batting_new$current_pred
rmse_valid[1] <- rmse(batting_valid, truth = salary, estimate = current_pred) |>
  pull(.estimate)
rmse_valid[1]
```

To absolutely no one's surprise, this is an absolutely terrible model.

## Step 2: Fit the Best Model on the Residuals

The first trick to slow learning is that we always want to fit the next model to predict the *prediction errors* from the previous model. You can think of this as trying to learn where and why the previous model screwed up.

### Step 2a: Find the Best Model

We're going to fit trees that can only split based on 1 variable, because this lesson isn't going to bother doing cross-validation to figure out the right interaction depth.

```{r}
#| label: step 2a setup
predictor_possibilities <- c(
  "G", "R", "H", "X2B", "X3B", "HR", "RBI",
  "SB", "CS", "BB", "SO", "IBB", "HBP", "SH", 
  "SF", "GIDP","age", "lgID"
)
```

We search over all predictors to see which model produces the lowest validation set RSS.

```{r}
#| label: step 2a search over predictors

p <- length(predictor_possibilities)

RSS <- numeric(p)     # p 0's

for (j in seq_along(predictor_possibilities)) {
  candidate_formula <- paste("current_resid ~ ",
                             predictor_possibilities[j])
  candidate_model <- rpart(
    formula = as.formula(candidate_formula),
    data = batting_new,
    maxdepth = 1     # not sure if this is right, but we only have one variable in our formula anyway
  )
  valid_predictions <- predict(candidate_model,
                               newdata = batting_valid)
  candidate_residuals <- batting_valid$current_resid - valid_predictions
  # Observed resid - predicted resid from previous model
  RSS[j] <- sum(candidate_residuals^2)
}

best_predictor <- predictor_possibilities[which.min(RSS)]
# find the index corresponding to the minimum value in RSS

best_predictor

```

It looks like a tree with a split based on IBB produces the minimum RSS.

### Step 2b: Obtain the New Predicted Residuals

Now we re-fit the model using our best predictor.

```{r}
#| label: Step 2b fit model

final_formula <- paste("current_resid ~ ", best_predictor)

fhat[[1]] <- rpart(
  formula = as.formula(final_formula),
  data = batting_new
)

```

```{r}
#| label: Step 2b make predictions

yhat1 <- predict(fhat[[1]], batting_new)

yhat_valid <- predict(fhat[[1]], newdata = batting_valid)

```

## Step 3: Update the Model

We only add in a *fraction* of the predictions, thus updating only a *fraction* of the residuals. This fraction is controlled by the learning rate $\lambda$.

The larger $\lambda$ is, the more aggressively the model adjusts its predictions. This generally achieves convergence in fewer iterations, but can overcorrect. A small value of $\lambda$ adjusts the predictions much more conservatively, but it may take longer to fit. In a real model, we would want to tune the learning rate, but again, this is a didactic exercise; let's just hard-code $\lambda = 0.1$, that is, add in only 10% of the prediction on every iteration.

```{r}
#| label: step 3 update model

lambda <- .1

batting_new <- batting_new |>
  mutate( # overwrite the training set
    current_pred = current_pred + lambda * yhat1,
    current_resid = current_resid - lambda * yhat1
  )

batting_valid <- batting_valid |>
  mutate( # overwrite the training set
    current_pred = current_pred + lambda * yhat_valid,
    current_resid = current_resid - lambda * yhat_valid
  )

salary_predictions[[2]] <- batting_new$current_pred
ggplot(data = batting_valid,
       mapping = aes(x = salary, y = current_pred)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1) +
  labs(x = "Actual Salary", y = "Predicted Salary (Step 1)")

rmse_valid[2] <- rmse(batting_valid, truth = salary, estimate = current_pred) |>
  pull(.estimate)

rmse_valid[2]

```

Notice that our validation set RMSE has decreased by only about 0.02. That's by design - this is only the first iteration. On each iteration, we intend to decrease the validation set RMSE.

## Step 4: Repeat

Each time through the loop, we have to do Steps 2 and 3:

```{r}
#| label: step 4 repeat
# outer for loop - each iteration
for (stepno in 2:niterations) {
  RSS <- numeric(p)     # reset the RSS vector

# inner for loop - searching over each candidate model
  
  for (j in seq_along(predictor_possibilities)) {
  candidate_formula <- paste("current_resid ~ ",
                             predictor_possibilities[j])
  candidate_model <- rpart(
    formula = as.formula(candidate_formula),
    data = batting_new,
    maxdepth = 1     # not sure if this is right, but we only have one variable in our formula anyway
  )
  valid_predictions <- predict(candidate_model,
                               newdata = batting_valid)
  candidate_residuals <- batting_valid$current_resid - valid_predictions
  # Observed resid - predicted resid from previous model
  RSS[j] <- sum(candidate_residuals^2)
  }
  
  best_predictor <- predictor_possibilities[which.min(RSS)]
  
  final_formula <- paste("current_resid ~ ", best_predictor)
  
  fhat[[stepno]] <- rpart(
  formula = as.formula(final_formula),
  data = batting_new)
  
  yhat1 <- predict(fhat[[stepno]], batting_new)
  
  yhat_valid <- predict(fhat[[stepno]], newdata = batting_valid)
  
  batting_new <- batting_new |>
  mutate( # overwrite the training set
    current_pred = current_pred + lambda * yhat1,
    current_resid = current_resid - lambda * yhat1
  )

batting_valid <- batting_valid |>
  mutate( # overwrite the training set
    current_pred = current_pred + lambda * yhat_valid,
    current_resid = current_resid - lambda * yhat_valid
  )

  current_rmse <- rmse(batting_valid,
                       truth = salary,
                       estimate = current_pred) |>
    pull(.estimate)
  
  cat(
    paste0(
      "Step No. ", stepno,
      ": Best Predictor = ", best_predictor,
      ", RMSE = ", round(current_rmse, 3), "\n"
    )
  )
  
  salary_predictions[[(stepno + 1)]] <- batting_new$current_pred
  
  rmse_valid[(stepno+1)] <- current_rmse

}
```

```{r}
#| label: plot RMSE

data.frame(
  iter = 0:50,
  rmse = rmse_valid
) |>
  ggplot(
    aes(x = iter, y = rmse)
  ) +
  geom_point() +
  geom_line() +
  labs(x = "Iteration Number", y = "Validation RMSE")

```


As you can see, the RMSE drops quickly and then starts leveling off, but it hasn't yet hit a minimum. Realistically, we don't want to train all the way down to the minimum RMSE, because that will almost certainly overfit. (Deep neural networks are a different story; a major assumption of deep learning is that there is so little noise in the data compared to the true complicated $f(x)$ that training down to 0 RMSE is actually desired.)

Let's see how a few of our predictions panned out:

```{r}
#| label: Albert Pujols
iter_pred <- as.data.frame(salary_predictions)
iter_pred <- iter_pred |>
  mutate(salary = batting_new$salary,
         name = paste(batting_new$nameFirst, batting_new$nameLast))


# Albert Pujols
plot(0:niterations, (iter_pred |> filter(name == "Albert Pujols"))[,1:(niterations+1)],
     xlab = "Iteration Number", ylab = "Predicted Salary",
     ylim = c(6.4, 7.5),
     main = "Albert Pujols")
abline(h = (iter_pred |> filter(name == "Albert Pujols"))$salary)
```

Albert Pujols exemplifies what we *want* the model to be doing. We start out at a very bad prediction and slowly get better on most iterations, even though we don't ever quite get to the true salary.

```{r}
#| label: Mike Trout
# Mike Trout
plot(0:niterations, (iter_pred |> filter(name == "Mike Trout"))[,1:(niterations+1)],
     xlab = "Iteration Number", ylab = "Predicted Value", ylim = c(6.1, 7.3),
     main = "Mike Trout")
abline(h = (iter_pred |> filter(name == "Mike Trout"))$salary)
```

This is what we *don't* expect to see. The model starts moving in the wrong direction and never really recovers. However, remember that we are minimizing the validation RSS, so making worse predictions for Trout may be seriously offset by improvements in predictions for players in the validation set.

```{r}
#| label: Jose Altuve
# Jose Altuve
plot(0:niterations, (iter_pred |> filter(name == "Jose Altuve"))[,1:(niterations+1)],
     xlab = "Iteration Number", ylab = "Predicted Salary", ylim = c(6.2, 6.6),
     main = "Jose Altuve")
abline(h = (iter_pred |> filter(name == "Jose Altuve"))$salary)
```

Sometimes you get patterns like this - we start off going in the wrong direction, but then eventually start figuring it out and going back toward the actual response value.

## Step 5: Make Predictions

Since we overwrote `fhat` on every iteration, we don't know at what values any of the splits occurred, so we can't actually make predictions on the test set with this boosted tree. 

```{r}
#| label: make predictions

batting_test <- batting_test |>
  mutate(
    .pred_salary = mean(batting_train$salary)
  )

for (stepno in 1:niterations) {
  .pred_step <- predict(fhat[[stepno]], newdata = batting_test)
  
  batting_test <- batting_test |>
    mutate(
      .pred_salary = .pred_salary + lambda *.pred_step
    )
}

ggplot(
  data = batting_test,
  mapping = aes(x = salary, y = .pred_salary)
) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0) +     # line of perfect calibration
  labs(
    x = "Actual Salary",
    y = "Predicted Salary (Test Set)"
  )

rmse(batting_test, truth = salary, estimate = .pred_salary)

```

It perhaps shouldn't surprise us that without doing literally any cross-validation, this RMSE isn't great. But this was just to illustrate the concept and basic algorithmic approach of slow learning.