---
title: "Simulation Activity: Two-Sample t-Test"
subtitle: "Math 437"
format: html
editor: source
---

## Goals

1. Simulate data from a normal distribution in R
2. Perform a two-sample t-test in R
3. Understand the difference between data in "wide format" and "long format"
4. Understand the `htest` object produced by most hypothesis tests in R

## Simulate the Data

We will start by creating *vectors* in R. A vector is basically just a list of numbers (or characters, or other data types).

The chunk below creates a variable named `group1` and assigns to it a vector consisting of 50 random numbers from a normal distribution with mean 100 and standard deviation 15. Similarly, the variable `group2` is assigned a vector consisting of 50 random numbers from a normal distribution with mean 90 and standard deviation 15.

```{r}
#| label: simulate data

set.seed(117)

group1 <- rnorm(
  50,                         # number of random numbers to generate
  mean = 100,                 # population mean
  sd = 15                     # population sd
)

group2 <- rnorm(
  50,                         # number of random numbers to generate
  mean = 90,                 # population mean
  sd = 15                     # population sd
)

```

## T-test with vectors

We can pass the two vectors directly into the `t.test` function to determine whether the population means of `group1` and `group2` are different.

```{r}
#| label: run a t-test with vectors

t.test(group1, group2)

```

Because we did not store the output as a vector, R prints out what it thinks we need to know. Sometimes this is the output itself, and sometimes (as in this case) it's a summary of the output. To actually use the output in later parts of the code, we need to store the output in a variable. In the chunk below, we rerun the test, but now store the output in a variable called `t_test_sim`:

```{r}
#| label: store output of the t-test

t_test_sim <- t.test(group1, group2)

```

Notice that no output is produced. If we want to print the output, it is easiest to add another line of code to explicitly tell R to `print` the variable, as shown in the chunk below.

```{r}
#| label: print output of the t-test

print(t_test_sim)

```


### Investigating the htest object

The variable `t_test_sim` now appears in your Environment tab in RStudio as a "List of 10." In R, the output of a hypothesis test is an `htest` object containing information about the methods and results of the inference. Let's see what information we can get out of the `t_test_sim` object.

`htest` is a special class of `list`, an R object that contains one or more other R objects. Any time we have a list-like object, we can use the `names` function to identify what the objects in the list are named.

```{r}
#| label: investigate contents of t_test_sim

names(t_test_sim)

```

We can extract the contents of a specific object within the list using the `$` operator. For example, the code below extracts the p-value for the test:

```{r}
#| label: get p-value out

t_test_sim$p.value

```

## Wide vs. Long Format

When vectors have the same length, we can combine them into a data structure called a *data frame*. Most datasets you import into R will by default be either a data frame or a special type of data frame called a *tibble*, depending on how you import the data.

The code below creates a data frame named `sim_df_wide` out of the two vectors `group1` and `group2`.

```{r}
#| label: create a data frame

sim_df_wide <- data.frame(
  group1_data = group1, 
  group2_data = group2
)

```

This data frame is considered to be in "wide" format, where different columns represent the values of the same variable within different groups. The columns of the data frame are treated equivalently to objects within a list. Consider what happens when we use the `names` function on our data frame:

```{r}
#| label: names on a data frame

names(sim_df_wide)

```

This means that we can use the `$` operator to extract the contents of that column as a vector and pass it into the `t.test` function:

```{r}
#| label: t-test on columns of a data frame

t.test(
  sim_df_wide$group1_data,
  sim_df_wide$group2_data
)

```


However, a more typical data representation has different rows represent data collected from different observational units and each column represent a unique variable whose values are being recorded. Thus, each number/string/etc. in the data frame represents the value of a specific variable (determined by the column it's in) recorded from a specific observational unit (determined by the row it's in). This data format is sometimes referred to as "long" format and is what the `tidyverse` ethos refers to as "tidy data."

The code below constructs a data frame in long format by explicitly creating a *character* vector to contain the group names, combining all of the numbers into a single numeric vector, and then using the `data.frame` function to create the data frame out of those two vectors. There are functions in the `tidyverse` packages that convert between "wide" and "long" formats, but doing it this way helps us understand the distinction between the two formats.

```{r}
#| label: long format data

group <- c(
  rep("Group 1", 50),
  rep("Group 2", 50)
)

sim_df_long <- data.frame(
  group = group,           # group on left is name of column, group on right is the vector in Environment whose values are being used
  value = c(group1, group2)
)

```

Notice that the same data is contained in both `sim_df_wide` and `sim_df_long`, but the shape of the data frame is different:

```{r}
#| label: dimensions of sim_df_wide and sim_df_long

dim(sim_df_wide)
dim(sim_df_long)

```

### Your Turn

Write code in the chunk below that finds the names of the variables in the data frame.

```{r}
#| label: variable names in sim_df_long

names(sim_df_long)

```

## T-test with formulas

Notice that all of the numerical values are now in the same column. While we could technically pass in a specific subset of the that vector as `x` and a different subset as `y`, there is a much easier way to use the `t.test` function when the data is in long format.

We instead pass in a `formula` argument of the form `response ~ explanatory`, where `response` and `explanatory` are the names of the response and explanatory variables. If those variables are (as usual) columns of a data frame, we also need a `data` argument to tell R what data frame to look in.

```{r}
#| label: t-test in long format

t_test_sim_long <- t.test(
  value ~ group,          # LHS is response variable, RHS is explanatory variable
  data = sim_df_long
)

print(t_test_sim_long)

```

### Your Turn

Write a chunk to extract the p-value from `t_test_sim_long`. Do you get the same p-value that you got originally?

```{r}
#| label: t-test in long format

t_test_sim_long$p.value

```

## Further Exploration

What information is contained the `statistic`,  `parameter`, and `conf.int` objects that are output by the `t.test` function? 

```{r}
#| label: extract statistic

t_test_sim_long$statistic

```

```{r}
#| label: extract parameter

t_test_sim_long$parameter

```


```{r}
#| label: extract confidence interval

t_test_sim_long$conf.int

```









