---
title: "Simulation Activity: Multiple Comparisons"
subtitle: "Math 437"
format: html
editor: source
---

## Goals

1. Estimate the Type I Error rate and power of a test
2. Understand the difference between Family-Wise Error Rate and False Discovery Rate
3. Investigate how these properties change depending on sample size and population distribution

## Simulate the Data

```{r}
#| label: simulate the data

n_tests <- 1000
sample_size <- 10
prop_true_H0 <- 0.5

n_true_H0 <- floor(n_tests*prop_true_H0)
n_false_H0 <- n_tests - n_true_H0

mu0 <- 0
sigma0 <- 1

mu1 <- 1
sigma1 <- 1

set.seed(6)
x_H0 <- matrix(
  rnorm(sample_size*n_true_H0, mean = mu0, sd = sigma0),
  nrow = sample_size,
  ncol = n_true_H0
  )
x_H1 <- matrix(
  rnorm(sample_size*n_false_H0, mean = mu1, sd = sigma1),
  nrow = sample_size,
  ncol = n_false_H0
  )
x <- cbind(x_H0, x_H1)
```

Add comments to the chunk above indicating what the code is doing. What does the final matrix `x` contain?

*19-21, sets values, 21 is the how many H0 is actually gonna be true
*23 & 24 gives us true and false H0, floor forces an integer
*mu and sigmas are properties of dist. under H0 and H(alternative) = Ha
*Seed is seed
*x_H0 builds a matrix, rnorm generates random numbers based on distribution, with amount based off true H0, mean, and sd.
Row is each obs, column is the true grouping?
*x_H1 is the same but false
*cbind does idk

## A Single t-Test

```{r}
#| label: run a one-sample t-test on the first column of x

t.test(x[,1], mu = mu0)

```

Do you reject the null hypothesis at significance level $\alpha = 0.05$? Why or why not?

## Multiple t-Tests

Three options here:

1. Inside a `for` loop, run a t-test on each column and store the corresponding p-value in the corresponding index of a `p_values` vector.
2. Write a function, `extract_p_value()`, to perform a t-test and extract the p-value, then `apply` the function over the columns of `x` and store the output in `p_values`.
3. As Option 2, but using an anonymous (lambda) function instead of defining the function `extract_p_value` directly.

```{r}
#| label: multiple t-tests

extract_p_value <- function(x, mu){
  # x : a vector of data
  # mu : the hypothesized value of mu under H0
  return(t.test(x, mu = mu)$p.value)    # explicitly return p-value
}

p_values <- apply(
  x,                 # matrix or data frame with the data
  2,                 # dimension to apply over
  extract_p_value,   # function to apply
  mu = mu0           # other arguments to pass to the function
  
)

```


```{r}
#| label: is H0 true

H0_status <- c(
  rep("H0 True", n_true_H0),
  rep("H0 False", n_false_H0)
)

alpha <- 0.05

H0_rejected <- ifelse(
  p_values <= alpha,     # condition to evaluate
  "Reject H0",           # what to return if TRUE : p-value <= alpha
  "Fail to reject H0"    # what to return if FALSE: p-value > alpha
)

table(H0_rejected, H0_status)

```

### Food for Thought

Estimate the following based on the table above.

1. The probability of committing a Type I Error on an individual test
    28/1000 or .028 or .056 based on just column
2. The probability of committing any Type I Errors
    1 (probability of committing any, as in at all, which is 100%)
3. The power of the test
    Power = P(reject H0 | some specific H1 true (mu = 1) )
    About .8 (not sure how)
4. The probability that a rejected null hypothesis will be a Type I Error
    Power = P(Type 1 Error | reject H0 ) = 28/(407+28) = .06
    False discovery Proportion = .064
    (estimate of FDR)

## Bonferroni Method

The Bonferroni Method adjusts $\alpha$ by dividing by the total number of tests.

```{r}
#| label: Bonferroni Method

alpha_bonf <- alpha/n_tests

H0_status <- c(
  rep("H0 True", n_true_H0),
  rep("H0 False", n_false_H0)
)

alpha <- 0.05

H0_rejected <- ifelse(
  p_values <= alpha_bonf, # condition to evaluate
  "Reject H0",            # what to return if TRUE : p-value <= alpha
  "Fail to reject H0"     # what to return if FALSE: p-value > alpha
)

table(H0_rejected, H0_status)

```

Did we make any Type I Errors? What proportion of false null hypotheses were rejected?

## Holm's Step-Down Method

Holm's Step-Down Method compares the p-values sequentially to increasing significance levels, and stops as soon as one p-value is found to be non-significant. Let's see how this works.

```{r}
#| label: holm method

# Step 1: Define the desired FWER
FWER <- .05
# Step 2: Order the p-values from smallest to largest
p_values_sorted <- sort(p_values, decreasing = FALSE)
# Step 3: Test the p-values at increasing significance levels until one is non-significant
holm_decisions <- rep("Fail to Reject H0", n_tests)

for(j in 1:n_tests){
  # Compare the j^th ordered p-value to the j^th ordered sig. level
  new_sig_level <- FWER/(n_tests + 1 - j)
  if(p_values_sorted[j] <= new_sig_level){
    # Find the correct null hypothesis to reject
    holm_decisions[which(p_values == p_values_sorted[j])] <- "Reject H0"
  } else {
    # If not rejected, STOP and do not reject any other H0
    break
  }
}

table(holm_decisions, H0_status)

```

Did we make any Type I Errors? What proportion of false null hypotheses were rejected?
No, 487/413 false null hypotheses were rejected

### Food for Thought

Does it really matter whether we use Holm's method or the Bonferroni method when we have a truly large (here 1000) number of tests? Why or why not?

## False Discovery Rate

Let's see if the Benjamini-Hochberg method can do any better. The Benjamini-Hochberg method gives up on controlling the family-wise error rate and instead tries to control the false discovery rate.

```{r}
#| label: B-H method

# Step 1: Define the desired FWER
FDR <- .10                       # No consensus on what to use, 10% and 20% are common
# Step 2: Order the p-values from smallest to largest
p_values_sorted <- sort(p_values, decreasing = FALSE)
# Step 3: Test the p-values at increasing significance levels until one is non-significant
BH_decisions <- rep("Fail to Reject H0", n_tests)

for(j in 1:n_tests){
  # Compare the j^th ordered p-value to the j^th ordered sig. level
  new_sig_level <- FDR*j/n_tests
  if(p_values_sorted[j] <= new_sig_level){
    # Find the correct null hypothesis to reject
    BH_decisions[which(p_values == p_values_sorted[j])] <- "Reject H0"
  } else {
    # If not rejected, STOP and do not reject any other H0
    break
  }
}

table(BH_decisions, H0_status)

```

Did we make any Type I Errors? If so, what is the false discovery proportion? What proportion of false null hypotheses were rejected?

## Further Exploration

Copy all of the code to simulate the data, run the t-tests, and apply corrections to control the FWER or FDR. With a partner or two, change the situation under which we are performing the tests:

1. Change the value of `mu0` and/or `mu1` to make the population distributions closer or further apart
2. Change the value of `sigma0` and/or `sigma1` to either change the variability in both population distributions or give them different variability
3. Change the distribution entirely, e.g., using `rexp` to simulate from an exponential distribution or `rpois` to simulate from a Poisson distribution.
4. Change the sample size - to make the code and principles work we need to keep this the same under both $H_0$ and $H_1$
5. Change the proportion of samples for which $H_0$ is true

```{r}
#| label: simulate data again


```

```{r}
#| label: Bonferroni method on new simulated data

```

```{r}
#| label: Holm method on new simulated data

```

```{r}
#| label: B-H method on new simulated data

```

What do you notice about the effect of your modification? If you have time, try a couple of different modifications. 