---
title: "Simulation Activity: Bootstrap Confidence Intervals"
subtitle: "Math 437"
format: html
editor: source
---

## General Format of a Bootstrap Confidence Interval

Remember that the confidence level describes a *procedure* for generating the interval from a sample of data. The procedure that we use to create a bootstrap confidence interval is:

1. Generate a bunch of bootstrap samples
2. Calculate something on each bootstrap sample
3. Use the calculated numbers from Step 2 to obtain the confidence interval

The distinction between the different bootstrap confidence interval methods is what happens in Step 2 and Step 3.

Let's try some different bootstrap methods for estimating a population mean. First we'll simulate some data so that we know what the "right" population mean is:

```{r}
#| label: simulate our data

set.seed(138)
x <- rexp(100, rate = 1)  # we know that the pop. mean = 1
hist(x)                   # this looks sketch

```

## Bootstrap Percentile Confidence Interval

This is the simplest method for getting a confidence interval. 

Step 2: Calculate an estimate of the parameter, producing a bootstrap sampling distribution of an appropriate estimator.

Step 3: Find the $\alpha/2$ and $1 - \alpha/2$ percentiles of the bootstrap sampling distribution.

```{r}
#| label: construct bootstrap sampling distribution

library(boot)

B <- 1000

boot_sample_mean <- function(df, ind){
  # df = data frame containing the original data to sample from
  # ind = vector of row indices corresponding to the bootstrap sample
  
  # Step 1: the bootstrap sample is inherent when we subset df based on ind
  boot_df <- df[ind, , drop = FALSE]
  
  # Step 2: calculate the statistic
  boot_mean <- mean(boot_df$x)  # Note to self: we better have a column labeled "x"
  
  return(boot_mean) # explicitly what is the output
  
}

sim_data <- data.frame(x=x)

set.seed(142)
boot_sampling_dist <- boot(
  data = sim_data,
  statistic = boot_sample_mean,
  R = B
)

```

```{r}
#| label: bootstrap percentile interval

hist(boot_sampling_dist$t)
C <- .95                    # confidence level as a decimal
alpha <- (1-C)              # sig. level + conf. level = 1
percentile_bounds <- quantile(
  boot_sampling_dist$t,     # vector we are finding percentiles of
  c(alpha/2, 1-alpha/2)     # vector of percentiles to use
)

print(percentile_bounds)

```

Three nice properties of a bootstrap percentile interval

1. Intuitive (no need to mess around with what critical value)
2. Range-Perserving (guarentees no ridiculous estimates)
3. Transformation-invariant: if $\phi = m(\theta)$, where m is a monotonic transformation, and the CI for $\theta$ is (L, U), then the CI for $\phi$ is $m((L), m(U))$ or $m((U), m(L))$ if m is a decreasing fn

### Simulation Study of the Confidence Level

Recall that the confidence level describes the probability that the method will "work" prior to collecting data, that is, the long-run proportion of samples for which our method will produce an interval that contains the true value of the parameter to be estimated. (Technically, there's an $\text{inf}$ involved, but we're going to ignore the weird edge cases...)

This means that we can estimate the confidence level by:

1. Obtaining a very large number of samples
2. Computing a confidence interval on each sample
3. Determining the proportion of the resulting confidence intervals that contain the true parameter value

Obviously, we can only do this using simulation. So let's do it.

```{r}
#| label: turn percentile CI into a function

boot_percentile_ci <- function(x, B = 1000, C = 0.95){
  # this function constructs a single bootstrap percentile CI
  # x = a vector of data
  # B = the number of bootstrap samples
  # C = the confidence level
  
sim_data <- data.frame(x = x)
  
  # get sampling dist.
boot_sampling_dist <- boot(
  data = sim_data,
  statistic = boot_sample_mean,
  R = B
)

  # get quantiles of the sampling dist.
  alpha <- (1-C)              # sig. level + conf. level = 1
  percentile_bounds <- quantile(
  boot_sampling_dist$t,     # vector we are finding percentiles of
  c(alpha/2, 1-alpha/2)     # vector of percentiles to use
)

return(percentile_bounds)
}
```

```{r}
#| label: lots of simdata

lots_of_simdata <- matrix(
  rexp(1000*100, rate = 1),
  nrow = 100,
  ncol = 1000
)

```

```{r}
#| label: percentile CI on each random sample
# now we need to apply our function over each column of simulated data


set.seed(206)
boot_perc_cis <- apply(
  lots_of_simdata,      # matrix w/ data in it
  2,                    # apply over each COLUMN in the matrix
  boot_percentile_ci    #function to apply
)

```

```{r}
#| label: estimate the confidence level

boot_low <- boot_perc_cis[1, ]
boot_high <- boot_perc_cis[2, ]
cover <- boot_low < 1 & boot_high > 1 # TRUE if 1 is between boot_low and boot_high, FALSE o/w
mean(cover)
# this gives proportion of our 1000 CIs that cover 1
```

## Bootstrap Bias-Corrected and Accelerated (BCa) Confidence Interval

In Step 3, the BCa method "adjusts" the bounds of the percentile confidence interval to (usually) preserve its properties while minimizing the disadvantages, at the cost of additional computation.

```{r}
#| label: manually do the BCa adjustments
# start with bias correction
observed_estimate <- mean(simdata)
prop_below_orig <- sum(bootstrap_estimates <  observed_estimate)/B
z0.hat <- qnorm(prop_below_orig)

# acceleration requires jackknifing
jackknife_estimates <- numeric(n)  # we remove each observation once
for (i in 1:n){
  newdata <- simdata[-i]  # remove the ith observation from the vector
  jackknife_estimates[i] <- mean(newdata)  # compute the mean of all but the ith observation
}

jackknife_mean <- mean(jackknife_estimates)

a.temp <- jackknife_mean - jackknife_estimates
a.numerator <- sum(a.temp^3)
a.denominator <- 6*(sum(a.temp^2))^(3/2)
a.hat <- a.numerator/a.denominator

# compute the corrections
fraction.low <- (z0.hat + qnorm(alpha/2))/(1 - a.hat*(z0.hat + qnorm(alpha/2)))
fraction.high <- (z0.hat + qnorm(1-alpha/2))/(1 - a.hat*(z0.hat + qnorm(1-alpha/2)))

# compute the actual quantiles based on the corrections
BCa.quantiles <- pnorm(c(fraction.low, fraction.high))

# compute the confidence interval
boot_bca_ci <- quantile(bootstrap_estimates, probs = BCa.quantiles)
print(boot_bca_ci)
```

Notice that our estimates were moved slightly higher because the sample was skewed right. 

## Constructing Bootstrap Confidence Intervals Using `boot`

```{r}
#| label: load packages
library(boot)
```

Here we reverse Step 1 and Step 2. First we define a function that we're going to apply to every bootstrap sample, then we create the bootstrap samples with `boot`.

```{r}
#| label: Step 1 define the function

boot_sample_mean <- function(df, ind){
  # df = data frame containing the original data to sample from
  # ind = vector of row indices corresponding to the bootstrap sample
  
  # Step 1: the bootstrap sample is inherent when we subset df based on ind
  boot_df <- df[ind, , drop = FALSE]
  
  # Step 2: calculate the statistic
  boot_mean <- mean(boot_df$x)  # Note to self: we better have a column labeled "x"
  
  return(boot_mean) # explicitly what is the output
  
}

```

```{r}
#| label: Step 2 create bootstrap samples
B <- 1000

sim_data <- data.frame(x=x)

set.seed(142)
boot_sampling_dist <- boot(
  data = sim_data,
  statistic = boot_sample_mean,
  R = B
)

```

Finally, in Step 3, we use the `boot.ci` function to construct the confidence interval(s):

```{r}
#| label: Step 3 with boot.ci

boot.ci(
  boot_sampling_dist, # the thing we got out of the boot fn
  conf = .95,         # conf int
  type = c("perc", # percentile
           "bca"   # BCa CI
           )
)

```

Note that these intervals differ slightly because the bootstrap samples are different.

