---
title: "Simulation Activities: Indicator Variables and Interaction Effects"
subtitle: "Math 437"
format: html
editor: source
---

## Simulating with Indicator Variables

Let's simulate this model using one numerical and one categorical predictor:

```{r}
#| label: simulate multiple linear regression

set.seed(141)
x<- rnorm(500, mean = 5, sd = 3)
group <- c(
  rep("A", 100),
  rep("B", 200),
  rep("C", 200)
)

## Consistent slope in all 3 groups
beta1 <- 2
## Different intercepts in all 3 groups
betaA <- -1
betaB <- 1
betaC <- 3

# Build the true model (group == "?") = 1 or 0
y <- betaA * (group == "A") + # intercept if group is A
     betaB * (group == "B") + # intercept if group is B
     betaC * (group == "C") + # intercept if group is C
     beta1 * x +              # incorporate effect of x in the model
     rnorm(500)               # random epsilon ~ N(0, 1)

simdata_indicators <- data.frame(
  y = y,
  x = x,
  group = group
)

```

### Explaining the Simulation

What is the true model relating x and y in Group A? Group B? Group C?

In Group A, y = -1 + 2x + epsilon
In Group B, y = 1 + 2x + epsilon
In Group C, y = 3 + 2x + epsilon

### Visualizing the Simulation

```{r}
#| label: visualize simulation

# subset our data to each group
A <- simdata_indicators[which(simdata_indicators$group == "A"), ]
B <- simdata_indicators[which(simdata_indicators$group == "B"), ]
C <- simdata_indicators[which(simdata_indicators$group == "C"), ]

# get range of x & y
xrange <- range(simdata_indicators$x) # min & max
yrange <- range(simdata_indicators$y)

plot(A$x, A$y, xlim = xrange, ylim = yrange, col = "green4")
# group A in blue
abline(a = betaA, b = beta1, col ="green4")
points(B$x, B$y, col = "red")
abline(a = betaB, b = beta1, col = "red")
points(C$x, C$y, col = "blue")
abline(a = betaC, b = beta1, col = "blue")


```


## Indicator Variables

Notice that `group` is categorical. We would like to create a single linear regression model that incorporates the effect of group, but we can't multiply a group value by a slope and get a number out. Instead, R changes the `group` variable so that we can include it in the model:

```{r}
#| label: indicator variable in model

lm_indicators <- lm(
  y ~ x + group,
  data = simdata_indicators
)

summary(lm_indicators)

```

### Interpreting the Output

1. What do the variables `groupB` and `groupC` mean?

groupB = 1 if group == "B", 0 otherwise
groupC = 1 if group == "C", 0 otherwise

groupA is defined by the result of them 2, any unrepresented indicator like groupA is referred to as the reference level or baseline.

2. What is the reference level (baseline) for `group` in this model? Why?

3. What is our least-squares regression equation?

yhat = -.987 + 1.997(x) + 1.803(group == "B") + 4.03(group == "C")

4. How do we interpret the slope corresponding to `groupB` in this model? What about `groupC`?

Slope was set as 1 for B, why do we have 1.803 now?
In groupA: yhat = -.987
In groupB: yhat = -.987 + 1.803
In groupC: yhat = -.987 + 4.03
Slope corresponding to indicator variable: Change in predicted y when we change groups from the ref. level to group of interest (holding other predictors constant)

### Changing the Reference Level

To use a different reference level, we need to use the `relevel` function. However, this function only works on factor variables:

```{r}
#| label: use relevel

simdata_indicators$group <- simdata_indicators$group |>
  as.factor() |> # convert to a factor variable
  relevel(ref = "B")

```

Notice what happens with the coefficients:

```{r}
#| label: compare coefficients

lm_refB <- lm(
  y ~ x + group,
  data = simdata_indicators
)

summary(lm_refB)

```

1. Explain how we could compute the intercept and slope for the new model using the coefficients from the original model.

## Interaction Effects

```{r}
#| label: simulate multiple linear regression with interaction

set.seed(100)
x1 <- rnorm(500, mean = 5, sd = 3)   # simulate from Normal
x2 <- rpois(500, lambda = 5)   # simulate from Poison
epsilon <- rnorm(500, mean = 0, sd = 5)

beta0 <- 2 # intercept
beta1 <- 3 # slope for x1
beta2 <- 7 # slope for x2
# at this point, mu_Y = 2 +3(X1) + 7(x2)
beta3 <- -1

y <- beta0 + beta1 * x1 + beta2 * x2 + beta3 * x1 * x2 + epsilon


```

### Explaining the Simulation

1. What is the difference between a *main effect* and an *interaction effect*?
all in notebook

2. How do we interpret `beta1` in this model? What about `beta2`?
beta1 is the change in mu given a 1 unit change in x1, holding x2 constant at 0. AND VICE VERSA

3. How do we interpret `beta3` in this model?
beta 3 represents the change in the slope of the regression line for y vs x1 when x2 increases by 1 AND VICE VERSA

## Fitting a Linear Regression Model with an Interaction Effect

An interaction effect is coded with a `:`

```{r}
#| label: interaction effect model 1

simdata_interaction <- data.frame(
  x1 = x1,
  x2 = x2,
  y = y
)

#this is the actual analysis done in lab
lm_no_interaction <- lm(y ~ x1 + x2,                # no interaction effect
                        data = simdata_interaction)
summary(lm_no_interaction)

lm_interaction <- lm(y ~ x1 + x2 + x1:x2,           # includes interaction effect between x1 and x2
                        data = simdata_interaction)
summary(lm_interaction)

```

### Interpreting the Output

1. What is our least-squares regression equation?
yhat = 4.58 + 2.50(x1) + 6.57(x2) + (-.91)(x1)(x2)

2. How do we interpret the slope corresponding to `x1` in this model?
When x1 increases by 1 unit, and x2 is held constant at 0, predicted y increases by 2.50 units.
As x2 increases, the slope connecting x1 & y decreases, at some x2 value it will be 0, and then the x1-y relationship will be negative.

3. What does it mean for the interaction term to have a negative slope?
Antagonistic interaction effect - "Diminishing Returns"
Synergistic Interaction Effect (If interaction slope>0) ("1+1=3")

### Including Main and Interaction Effects in the Same Model

```{r}
#| label: main and interaction effects

lm_interaction2 <- lm(
  y ~ x1 * x2, # include x1, x2, and their interaction
  data = simdata_interaction
)

summary(lm_interaction2)

```

## Quantitative-Qualitative Interactions

We can look at the combined effect of a quantitative and categorical predictor on a quantitative response by showing the different categories in different colors on a scatterplot.

```{r}
#| label: simulated group interaction model

set.seed(141)

x1 <- rnorm(500, mean = 5, sd = 3)   # simulate from Normal
group <- c(rep("A", 250), rep("B", 250))

# true slope and intercept in group A
beta0A <- 2
beta1A <- 3

# true slope and intercept in group B
beta0B <- -3
beta1B <- 5

y <- numeric(500)    # empty vector of zeros

# Group A
y[1:250] <- beta0A + beta1A * x1[1:250] + epsilon[1:250]

# Group b
y[251:500] <- beta0B + beta1B * x1[251:500] + epsilon[251:500]

simdata_group_interaction <- data.frame(
  x1 = x1,
  group = group,
  y = y
)

```

```{r}
#| label: plot without color

plot(y ~ x1, data = simdata_group_interaction)

```


```{r}
#| label: plot with color

colors <- c(
  rep("magenta4", 250), # group A
  rep("darkorange", 250)   # group B
)

plot(y ~ x1, data = simdata_group_interaction, col = colors)

```

Here's what the same plot looks like using the `ggplot2` package and adding the least-squares regression lines:

```{r}
#| label: interaction plot with ggplot2
#| warning: false
#| message: false

library(ggplot2)
ggplot( # set up the axes
  data = simdata_group_interaction,
  mapping = aes( # which variables map to which features of the plot
    x = x1,            # means put x1 on y-axis
    y = y,             # means put y on y-axis
    color = group      # means put different groups in different colors
    )
) +                    # add to this plot
  geom_point() +       # add points
  scale_color_manual(  # change the colors
    values = c(A = "magenta4", B = "darkorange")
  ) +
  geom_smooth(method = "lm", se = FALSE)

```

### Interpreting the Graphs

What in these graphs suggests that there is an interaction effect?

### Interpreting the Model

Here's the actual linear regression model:

```{r}
#| label: group interaction model

lm_group_interaction <- lm(
  y ~ x1 * group, # main effects of x1 and group and their interactions
  data = simdata_group_interaction
)

summary(lm_group_interaction)

```

1. How do we interpret the slopes corresponding to each main effect?
When x1 increases by 1 unit, in group A, predicted y increases by 3.11 units.
When group changes from A to B, and x1 is held constant at 0, predicted y decreases by 4.91 units.

2. How do we interpret the slopes corresponding to each interaction effect?
When group changes from A to B, the slope relating y &x1 increases by 1.89 units

3. What would be the equation for predicting y from x in group A? What about in group B? Group C? How close did those equations come to the true model?
yhat = 1.885 + 3.11(x1) - 4.91(group = "B") + 1.89(x1)(group = "B")
yhat = 1.885 + 3.11(x1)   (when group = A)
yhat = 1.885 + 3.11(x1) - 4.91 + 1.89(x1)   (when group =B)
    = -3.025 + 5(x1)

### Changing the Reference Level

```{r}
#| label: new reference level group interaction

simdata_group_interaction$group <- 
  as.factor(simdata_group_interaction$group) |>
  relevel(ref = "B")   # main effects of x1 and group and their interaction

lm_group_interaction2 <- lm(
  y ~ x1 * group, # main effects of x1 and group and their interactions
  data = simdata_group_interaction
)

summary(lm_group_interaction2)

```

Can you reconstruct the same equations for the three groups out of this model?

## Checking for Quantitative-Quantitative Interactions

The easiest way is to turn one of the quantitative variables into a categorical variable by *binning* it like you would in a histogram.

```{r}
#| label: bin x2

simdata_interaction$x2_group <- cut(
  x2,
  breaks = c(-1, 5, 10, 15)
)

table(simdata_interaction$x2_group)

```

```{r}
#| label: interaction plot with ggplot2 - 2

ggplot(
  data = simdata_interaction,
  mapping = aes(
    x = x1,
    y = y,
    color = x2_group
  )
) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)

```

