---
title: "Lab Assignment #5"
author: "Math 437 - Modern Data Analysis - Jason Soto"
format: pdf
editor: source
---

## Introduction

In this lab, we will fit logistic regression models in two ways: first, with the `glm` function, and then the `tidymodels` packages. Then, we will investigate measures of classification accuracy.

```{r}
#| label: do this at the beginning of every lab file
#| message: false
#| warning: false
here::i_am("Labs/Lab05.qmd")

# load packages
library(ISLR2)
library(nnet)
library(tidymodels)

# load data
cheating <- readr::read_csv(here::here("Data/cheating.csv"))
compas <- readr::read_csv(here::here("Data/compas.csv"))
```

This lab assignment is worth a total of **15 points**.

## Problem 1: Logistic Regression (4 pts total)

[Wilson and colleagues (2024)](https://www.sciencedirect.com/science/article/pii/S0022096524000882) investigated cheating behaviors in children on six different tasks The *cheating* dataset contains the age (in partial years) and gender (Boy or Girl) of each child tested, as well as two different response variables: whether they cheated on an impossible counting task (0 = No, 1 = Yes), and the total number of cheating behaviors observed across the six tasks.

### Part a (Code: 0.5 pts; "Explanation": 1 pt)

Using the `glm` function, create a logistic regression model predicting cheating on the counting task (`Cheated`) from Age and Gender. Write out the equation of the model (in either probability or log-odds form).

```{r}
#| label: Cheated Based on Age and Gender glm

cheat_model <- glm(
  Cheated ~ Age + Gender,
  data = cheating,
  family = "binomial"
)

summary(cheat_model)

```
In Log-Odds,
P(Cheating) = 2.4592 - .2956(Age) - .3471(Gender == Girl)

In Prob.,
P(Cheating) = (e^(2.4592 - .2956(Age) - .3471(Gender == Girl)))/(1 + e^(2.4592 - .2956(Age) - .3471(Gender == Girl)))


### Part b (Explanation: 1.5 points)

Interpret each number in the "Estimate" column of the summary created in Part (a).
According to the estimate column, in log-odds, with a y-intercept being 2.4592, a girl is less likely to cheat, and the chances of cheating decreases with age.

### Part c (Code: 1 point)

Use the `cheat_boy` and `cheat_girl` data frames below to predict the probability of cheating on the counting task.

```{r}
#| label: interpolation data frames

cheat_boy <- data.frame(
  Age = seq(4, 11.5, by = 1/12),
  Gender = "Boy"
)

cheat_girl <- data.frame(
  Age = seq(4, 11.5, by = 1/12),
  Gender = "Girl"
)

```

```{r}
#| label: predictions

boy_pred <- predict(
  cheat_model,
  newdata = cheat_boy,
  type = "response"
)

girl_pred <- predict(
  cheat_model,
  newdata = cheat_girl,
  type = "response"
)

```


Once you have your predictions, finish the chunk below to create a plot showing how the predicted probability of cheating changes with age for boys vs. girls. As the legend suggests, you should plot predicted probabilities for boys using black and for girls using red.

```{r}
#| label: cheating plot
#| fig-width: 6.5
#| fig-height: 5
# remove the eval: false line once you get this code to work!

plot(cheat_boy$Age,
     boy_pred,     # you need to put what you're plotting here
     type = "l",# type = "l" means make it a line
     xlab = "Age", ylab = "Predicted Prob. of Cheating",
     main = "Counting Task",
     xlim = c(4, 11.5), ylim = c(0,1),
     lwd = 2
     ) 
lines(cheat_girl$Age, girl_pred, col = "red", lwd = 2) # now add another solid line
legend("topright", legend = c("Boy", "Girl"), lty = c(1,1), col = c("black", "red"))
```

HINT: If your plot shows parallel lines instead of curves, you have forgotten a very important argument in your `glm` function in part (a)!

## Problem 2: Logistic Regression with tidymodels (11 pts total)

The "Correctional Offender Management Profiling for Alternative Sanctions" (COMPAS) is a proprietary algorithm used in several states (including California) to assess the risk that an arrested person will commit a(nother) crime after being released.

The tool has been heavily criticized not only for potential bias (especially against African-Americans), but also for its need to use a 137-question survey to achieve basically the same accuracy as much simpler models.

In this problem, we will use data collected by ProPublica from 2013 and 2014 arrests in Broward County, Florida and see how well we can predict an arrested person's future criminal activity from just three variables: their age, sex, and how many crimes they have previously been convicted of.

We will use the 2013 data as our training set and the 2014 data as our validation set.

```{r}
#| label: subset training vs. test (validation)

# factoring is done, no need to refactor
compas$recid <- factor(compas$recid,
                       levels = c("yes", "no"))

compas_train <- compas[which(compas$screening_year == 2013), ]
compas_test <- compas[which(compas$screening_year == 2014), ]
```

### Part a (Explanation: 2 pts)

Explain in the context of the COMPAS data what a true positive, true negative, false positive, and false negative would represent. You may assume that we are trying to predict whether an arrested person will be arrested again within two years of being released (`recid == "yes"`).

#### Solution

Positive class in this data set is the class that an arrested person is arrested again within two years of being released.

A True Positive here is that an arrested person is arrested again within two years of being released as predicted.
A True Negative here is that an arrested person isn't arrested again within two years of being released as predicted.
A False Positive here is that an arrested person isn't arrested again within two years of being released despite prediction of criminality.
A False Negative here is that an arrested person is arrested again within two years of being released despite prediction of innocence.

### Part b (Code: 1 pt)

Using the `tidymodels` packages, fit a logistic regression model on the `compas_train` dataset predicting `recid` from the three explanatory variables. Note that there may be some weirdness with figuring out the reference level for `recid`, so you may want to convert it to a factor variable and use the `relevel` function to set the reference level before fitting the model.

#### Solution

```{r}
#| label: Fitting

# Set up logistic regression model
compas_model <- logistic_reg(
  mode = "classification",
  engine = "glm"
)

# Set up recipe
compas_recipe <- recipe(
  recid ~ age_at_screening + sex + priors,
  data = compas_train
) |>
  step_dummy(all_nominal_predictors())

# Build workflow
compas_wflow <- workflow() |>
  add_model(compas_model) |>
  add_recipe(compas_recipe)

# Fit model
compas_fit <- fit(
  compas_wflow,
  data = compas_train
)

compas_fit

```


### Part c (Code: 1 pt)

Predict whether each arrested person in the `compas_test` dataset will be arrested within two years of release (`recid == yes`). Also, create a `compas_logr_predictions` data frame or tibble containing both the predicted and actual classes and the predicted probabilities of "yes" and "no". 

#### Solution

```{r}
#| label: fit and predict with logistic regression

compas_fit <- compas_wflow |>
  fit(data = compas_train)

compas_logr_predictions <- compas_fit |> 
  augment(new_data = compas_test)

compas_logr_predictions

```

### Part d (Code: 0.5 pts)

Use the `conf_mat` function to obtain the confusion matrix on the test set.

#### Solution

```{r}
#| label: Confusion Matrix

compas_logr_predictions$.pred_class <- as.factor(compas_logr_predictions$.pred_class)

conf.mat <- conf_mat(
  data = compas_logr_predictions,
  truth = recid, 
  estimate = .pred_class
)

conf.mat

```


### Part e (Code: 0.5 pts; Computation and/or Explanation: 2 pts)

Without running any additional code, use the confusion matrix to estimate the sensitivity (recall), specificity, positive predictive value (precision), and negative predictive value for the model. Express all answers as fractions and then convert to decimals rounded to the thousandths place (3 decimal places) or percentages rounded to the nearest tenth.

Using the `summary` function on your confusion matrix, check your answers.

#### Solution

Sensitivity = 25/(25+287) = .080
Specificity = 693/(693+11) = .984
Precision = 25/(25+11) = .694
NPV = 693/(693+287) = 707

```{r}
#| label: check

summary(conf.mat)

```

### Part f (Code: 2 pts)

Using the `roc_curve` function, obtain the ROC curve for this model. Also, using the `roc_auc` function, obtain the Area Under ROC Curve.

#### Solution

```{r}
#| label: ROC Curve
# Construct the ROC curve
roc_tibble <- roc_curve(
  data = compas_logr_predictions,
  truth = recid,
  .pred_yes     # Probability of positive class
)

# Plot the ROC curve

autoplot(roc_tibble)

```

```{r}
#| label: AUC

roc_auc(
  data = compas_logr_predictions,
  truth = recid,
  .pred_yes     # Probability of positive class
)

```

### Part g (Code: 0.5 pts)

Using the `brier_class` function, obtain the Brier score for this  model.

#### Solution

```{r}
#| label: Brier Class

brier_class(
  data = compas_logr_predictions,
  truth = recid,
  # must give it the prob. of first level
  .pred_yes
  # for K > 2 classes, need all .pred_{Class} probs.
)


```


### Part h (Code: 0.5 pts)

Using the `mn_log_loss` function, obtain the cross-entropy/log loss for this model.

#### Solution

```{r}
#| label: Log Loss

mn_log_loss(
  data = compas_logr_predictions,
  truth = recid,
  .pred_yes,     # prob of first level
  event_level = "first" # says we gave prob of 1st level in confusion matrix
)

```


### Part i (Explanation: 1 pt)

Give at least one reason why we might prefer to evaluate models based on their predicted probabilities instead of predicted classes. Give at least one reason why we might prefer to evaluate models based on their predicted classes instead of predicted probabilities.

#### Solution

Based on this model, we would prefer to evaluate models based on their predicted classes instead of predicted probabilities to simply categorize people into whether they will re-offend or not.
However, we could evaluate models based on their predicted probabilities instead of predicted classes to have a way to measure severity. For example a 52% chance to re-offend is much less dangerous to one who has a 99% chance to re-offend. This could be use to predict and solve prison vacancies issues.

