---
title: "Lab Assignment #8"
author: "Math 437 - Modern Data Analysis - Jason Soto"
format: pdf
editor: source
---

```{r}
#| label: do this at the beginning of every lab file
#| message: false
#| warning: false
here::i_am("Labs/Lab08.qmd")

library(ISLR2)
library(tidyverse)
library(broom)
library(recipes) # don't need the rest of tidymodels
library(naniar)
library(softImpute)

cereal3 <- readr::read_csv(here::here("Data/cereal3.csv"))
```

This lab assignment is worth a total of **15 points**.

## Problem 1: Principal Components Analysis

### Part a (Code: 1 pt)

Run the code in ISLR Lab 12.5.1. 

#### Solution

```{r}
#| label: 12.5.1 Chunk 1

states <- row.names(USArrests)
states

names(USArrests)

```

```{r}
#| label: 12.5.1 Chunk 2

apply(USArrests, 2, mean)

```

```{r}
#| label: 12.5.1 Chunk 3

pr.out <- prcomp(USArrests, scale = TRUE)

names(pr.out)

```

```{r}
#| label: 12.5.1 Chunk 4

pr.out$center
pr.out$scale
pr.out$rotation

```

```{r}
#| label: 12.5.1 Chunk 5

dim(pr.out$x)

```

```{r}
#| label: 12.5.1 Chunk 6

biplot(pr.out, scale = 0)

```

```{r}
#| label: 12.5.1 Chunk 7

pr.out$rotation = -pr.out$rotation
pr.out$x = -pr.out$x
biplot(pr.out, scale = 0)

```

```{r}
#| label: 12.5.1 Chunk 8

pr.out$sdev

```

```{r}
#| label: 12.5.1 Chunk 9

pr.var <- pr.out$sdev^2
pr.var

```

```{r}
#| label: 12.5.1 Chunk 10

pve <- pr.var / sum(pr.var)
pve

```

```{r}
#| label: 12.5.1 Chunk 11

par(mfrow = c(1, 2))

plot(pve,
     xlab = "Principal Component",
     ylab = "Proportion of Variance Explained",
     ylim = c(0, 1),
     type = "b")

plot(cumsum(pve),
     xlab = "Principal Component",
     ylab = "Cumulative Proportion of Variance Explained",
     ylim = c(0, 1),
     type = "b")

```

```{r}
#| label: 12.5.1 Chunk 12

a <- c(1, 2, 8, -3)
cumsum(a)

```

### Part b (Code: 1 pt)

The `cereal3` dataset contains nutrition and price information for 88 breakfast cereals that existed in 2019.

Using the `cereal3` dataset, create a new dataset called `cereal` that only contains the following variables:

* `Total.Fat`
* `Dietary.Fiber`
* `Sugar`
* `Protein`
* `Sodium`
* `Complex.Carbs`, which represents the carbohydrates that aren't dietary fiber or sugar

#### Solution

```{r}

cereal3$Complex.Carbs <- cereal3$Total.Carbohydrate - cereal3$Dietary.Fiber - cereal3$Sugar

cereal <- cereal3 |>
  select(
    Total.Fat, Dietary.Fiber, Sugar, Protein, Sodium, Complex.Carbs
  )

```


### Part c (Code: 1 pt)

Perform principal component analysis on the `cereal` dataset. You can use either the "base R" or "tidymodels" version of PCA. Remember to center and scale the data (either using the arguments `center = TRUE, scale = TRUE` in `prcomp` or using `step_normalize` before you do `step_pca`). 

#### Solution

```{r}
#| label: PCA

cereal_recipe <- recipe(
  ~ ., data = cereal
) |>
  step_normalize(all_numeric_predictors()) |>     # normalizes, converts to z-scores
  step_pca(all_predictors(), num_comp = 6)

cereal_prep <- prep(cereal_recipe)

cereal_baked <- cereal_prep |>
  bake(new_data = NULL)     # bake on the training set, no new data

cereal_tidy <- tidy(cereal_prep,
                 2,     # step 2 - PCA
                 type = "coef"     # extract loadings
                 )

head(cereal_tidy)

```


### Part d (Code: 2 pts; Explanation: 1 pt)

Find the proportion of variance explained by each component and the cumulative proportion of variance explained. Produce a scree plot showing either the proportion of variance explained or the cumulative proportion of variance explained. Suggest an appropriate number of principal components to use to visualize or interpret the data and justify your decision based on the scree plot.

#### Solution

```{r}

#| label: Proportion of Variance

cereal_pve <- tidy(
  cereal_prep,
  2,
  type = "variance"
)

cereal_pve

ggplot(
  data = cereal_pve |> filter(terms == "cumulative percent variance"),
  aes(x = component, y = value)
) +
  geom_point() +
  geom_line() +
  labs(
    x = "Principal Component",
    y = "Cumulative Proportion of Variance Explained"
  )

```
The cumulative scree plot shows how total variance increases as more principal components are included. The point where the curve begins to flatten shows that additional components contribute little additional variance.

### Part e (Code: 0.5 pts)

Produce a biplot showing the first two principal components.

#### Solution

```{r}
#| label: biplot on separate graphs

ggplot(
  data = cereal_baked,
  aes(x = PC1, y = PC2)) +
  geom_point() +
  labs(
    title = "Principal Component Scores",
       x = "PC1",
       y = "PC2")


cereal_loadings <- cereal_tidy |>
  pivot_wider(
    names_from = component,
    values_from = value
  ) |>
  dplyr::select(!id)


ggplot(
    data = cereal_loadings) +
  geom_segment(aes(
    x = 0, y = 0,
    xend = PC1, yend = PC2
  ),
  arrow = arrow(type = "open")
  ) +
  geom_text(
    aes(x = PC1,
        y = PC2,
        label = terms),
    check_overlap = TRUE
) +
  labs(
    title = "Principal Component Loadings",
     x = "PC1",
     y = "PC2")

```

### Part f (Explanation: 2 pts)

Give a word or short phrase describing what you think `PC1` represents about the cereals. Similarly, what do you think `PC2` represents? Explain your reasoning based on the biplot and/or the loadings matrix `rotation`.

#### Solution

PC1 may represent more high-protein and high-fiber cereals. We can deduce this as the protein, dietary fiber, and the complex carbs variables have the largest positive loadings. So PC1 is like old people cereal and stuff.

PC2 represents cereals that are higher in sugar and fat versus cereals that are lower in sugar and fat. We can deduce as the variables with the largest absolute values, sugar and total fat have the biggest loadings. So PC2 is more about differences in sugar and fat content.

## Problem 2: Missing Data

### Part a (Code: 1 pt)

Run the code in ISLR Lab 12.5.2. 

#### Solution

```{r}
#| label: 12.5.2 Chunk 1

X <- data.matrix(scale(USArrests))
pcob <- prcomp(X)
summary(pcob)

```

```{r}
#| label: 12.5.2 Chunk 2

sX <- svd(X)

names(sX)

round(sX$v, 3)

```

```{r}
#| label: 12.5.2 Chunk 3

pcob$rotation

```

```{r}
#| label: 12.5.2 Chunk 4

t(sX$d * t(sX$u))

pcob$x

```

```{r}
#| label: 12.5.2 Chunk 5

nomit <- 20
set.seed(15)

ina <- sample(seq(50), nomit)
inb <- sample(1:4, nomit, replace = TRUE)

Xna <- X

index.na <- cbind(ina, inb)

Xna[index.na] <- NA

```

```{r}
#| label: 12.5.2 Chunk 6

fit.svd <- function(X, M = 1) {
svdob <- svd(X)
with(svdob,
     u[, 1:M, drop = FALSE] %*% (d[1:M] * t(v[, 1:M, drop = FALSE]))
     )
}

```

```{r}
#| label: 12.5.2 Chunk 7

Xhat <- Xna

xbar <- colMeans(Xna, na.rm = TRUE)

Xhat[index.na] <- xbar[inb]

```

```{r}
#| label: 12.5.2 Chunk 8

thresh <- 1e-7

rel_err <- 1

iter <- 0

ismiss <- is.na(Xna)

mssold <- mean((scale(Xna, xbar, FALSE)[!ismiss])^2)

mss0 <- mean(Xna[!ismiss]^2)

```

```{r}
#| label: 12.5.2 Chunk 9

while(rel_err > thresh) {
  iter <- iter + 1
  # Step 2(a)
  Xapp <- fit.svd(Xhat, M = 1)
  # Step 2(b)
  Xhat[ismiss] <- Xapp[ismiss]
  # Step 2(c)
  mss <- mean(((Xna - Xapp)[!ismiss])^2)
  rel_err <- (mssold - mss) / mss0
  mssold <- mss
  cat("Iter:", iter, "MSS:", mss,
      "Rel. Err:", rel_err, "\n")
  }

```

```{r}
#| label: 12.5.2 Chunk 10

cor(Xapp[ismiss], X[ismiss])

```

### Part b (Code: 1 pt)

Adapt the code in ISLR Lab 12.5.2 to create a new data frame, `cereal_na_df`, which is exactly the same as the `cereal` data frame but missing 40 randomly removed data values. Each row should have at most one missing value.

#### Solution

Complete the code in this chunk to create `cereal_na_df`:

```{r}
#| label: cereal_missing

cereal_missing <- data.matrix(scale(cereal))

# Remove 40 values at random from cereal_missing 
set.seed(67)
missing_rows <- sample(1:nrow(cereal_missing), size = 40, replace = FALSE)
missing_cols <- sample(1:ncol(cereal_missing), size = 40, replace = TRUE)

cereal_na <- cereal_missing

# This assigns missing value
for(i in seq_along(missing_rows)){
  cereal_na[missing_rows[i], missing_cols[i]] <- NA
}

# Store the new matrix as cereal_na_df?
cereal_na_df <- as.data.frame(cereal_na)

```


### Part c (Code: 1 pt)

Produce a visualization showing the distribution of the missingness in `cereal_na_df`.

#### Solution

```{r}
#| label: Visualization

# default row and column order
vis_miss(cereal_na_df)

# orders the columns from most to least missing data
vis_miss(cereal_na_df, sort_miss = TRUE)

# reorder the rows so obs. of similar patterns of missingness are grouped
vis_miss(cereal_na_df, cluster = TRUE)

gg_miss_upset(cereal_na_df, nsets = 6)

```


### Part d (Explanation: 1.5 pts)

The lab suggests that we look into the `softImpute` package if we want to efficiently run the matrix completion code in Lab 12.5.2.  Run the following code to complete the matrix:

```{r}
#| label: use softImpute
cereal_svd <- softImpute(cereal_na)

## This line recreates the original matrix with NAs imputed
## I use softImpute::complete() because there is also a tidyr complete()
## that can hard-code missing values instead
cereal_completed <- softImpute::complete(cereal_na, cereal_svd)

```

Look at the `cereal_completed` matrix. Can you tell which values were removed and imputed? Why or why not?

#### Solution

No, we cannot tell which values were removed and imputed without looking at the matrix with the missing data. The imputed values do not stand out and look similar to the other values in the dataset. This is because they are estimated based on patterns in the data.

### Part e (Code: 1 pt; Explanation: 1 pt)

Produce a plot similar to Figure 12.5 in the textbook. You do not have to label or color-code the points, just plot the imputed values on the y-axis and the original values on the x-axis with the line $y = x$ superimposed.

Overall, does this algorithm appear to be doing a good job of imputing missing values? Does there appear to be any pattern in terms of which original values are over/under-estimated or which original values have more variation in the imputed values? Explain based on your plot.

#### Solution

```{r}
#|label: Plot

missing <- is.na(cereal_na)

original_values <- cereal_missing[missing]
imputed_values <- cereal_completed[missing]

plot(original_values, imputed_values,
     xlab = "Original Values",
     ylab = "Imputed Values"
     )

abline(0, 1, col = "red", lwd = 2)

```

Overall the algorithm looks alright. y = x implies that imputed values are equal to original values, and the data plots are pretty close to that line. We do have a few outliers but only a few. That is the best answer I can give, reading this plot is a little difficult.
