---
title: 'Lab Assignment #10'
author: "Math 437 - Modern Data Analysis - Jason Soto"
format: pdf
---

## Instructions

The purpose of this lab is to introduce several different classification strategies and variations on classification accuracy. In this lab we will work with, k-nearest neighbors, Naive Bayes and linear/quadratic discriminant analysis.

```{r}
#| label: load libraries and data
#| message: false
#| warning: false
here::i_am("Labs/Lab10.qmd")

# These are loaded in Lab 4.7
library(ISLR2)
library(MASS) # LDA/QDA
library(e1071) # Naive Bayes
library(class) # knn

# Might need these packages if you use tidyverse/tidymodels
library(tidyverse)
library(tidymodels)
library(discrim)
library(klaR)
library(kknn)

# compas data
compas <- readr::read_csv(here::here("Data/compas.csv"))
# factoring is done, no need to refactor
compas$recid <- factor(compas$recid,
                       levels = c("yes", "no"))
compas_train <- compas[which(compas$screening_year == 2013), ]
compas_test <- compas[which(compas$screening_year == 2014), ]
```


This lab assignment is worth a total of **17.5 points**.

For Problems 2-3, you may fit and predict using either the regular functions or the `tidymodels` workflows. Please do not mix and match. You should use functions from the `yardstick` package to do all model assessment.

## Problem 1: Understanding the Classification Algorithms

### Part a (Code: 4 pts)

Run the code in ISLR Labs 4.7.3-4.7.6. Put each chunk from the textbook in its own chunk. As usual, read the text to figure out what the chunks do, and don't `attach` anything (use the `with` function that we've learned about instead, e.g., `with(Smarket, plot(Volume))` instead of attaching Smarket).

```{r}
#| label: create train subset

train <- (Smarket$Year < 2005)
Smarket.2005 <- Smarket[!train, ]
Direction.2005 <- Smarket$Direction[!train]

```

### Section 4.7.3

(COMMENTED OUT PLOT, QUARTO WONT RENDER, SAYS figure margins too large)
```{r}
#| label: 4.7.3 Chunk 1

library(MASS)
lda.fit <- lda(Direction ~ Lag1 + Lag2,
               data = Smarket,
               subset = train)
lda.fit
#plot(lda.fit)

```

```{r}
#| label: 4.7.3 Chunk 2

lda.pred <- predict(lda.fit, Smarket.2005)
names(lda.pred)

```

```{r}
#| label: 4.7.3 Chunk 3

lda.class <- lda.pred$class
table(lda.class, Direction.2005)

mean(lda.class == Direction.2005)

```

```{r}
#| label: 4.7.3 Chunk 4

sum(lda.pred$posterior[, 1] >= .5)
sum(lda.pred$posterior[, 1] < .5)


```

```{r}
#| label: 4.7.3 Chunk 5

lda.pred$posterior[1:20, 1]
lda.class[1:20]

```

```{r}
#| label: 4.7.3 Chunk 6

sum(lda.pred$posterior[, 1] > .9)

```

### Section 4.7.4

```{r}
#| label: 4.7.4 Chunk 1

qda.fit <- qda(Direction ~ Lag1 + Lag2,
               data = Smarket,
               subset = train)

qda.fit


```

```{r}
#| label: 4.7.4 Chunk 2

qda.class <- predict(qda.fit, Smarket.2005)$class
table(qda.class, Direction.2005)

mean(qda.class == Direction.2005)

```

### Section 4.7.5

```{r}
#| label: 4.7.5 Chunk 1

library(e1071)

nb.fit <- naiveBayes(Direction ~ Lag1 + Lag2,
                       data = Smarket,
                       subset = train)
nb.fit

```

```{r}
#| label: 4.7.5 Chunk 2

mean(Smarket$Lag1[train][Smarket$Direction[train] == "Down"])

sd(Smarket$Lag1[train][Smarket$Direction[train] == "Down"])

```

```{r}
#| label: 4.7.5 Chunk 3

nb.class <- predict(nb.fit, Smarket.2005)
table(nb.class, Direction.2005)

mean(nb.class == Direction.2005)

```

```{r}
#| label: 4.7.5 Chunk 4

nb.preds <- predict(nb.fit, Smarket.2005, type = "raw")

nb.preds[1:5,]

```

### Section 4.7.6

```{r}
#| label: 4.7.6 Chunk 1

library(class)

train.X <- cbind(Smarket$Lag1, Smarket$Lag2)[train, ]

test.X <- cbind(Smarket$Lag1, Smarket$Lag2)[!train, ]

train.Direction <- Smarket$Direction[train]

```

```{r}
#| label: 4.7.6 Chunk 2

set.seed(1)
knn.pred <- knn(train.X, test.X, train.Direction, k = 1)

table(knn.pred, Direction.2005)

(83 + 43) / 252

```

```{r}
#| label: 4.7.6 Chunk 3

knn.pred <- knn(train.X, test.X, train.Direction, k = 3)

table(knn.pred, Direction.2005)

mean(knn.pred == Direction.2005)

```

```{r}
#| label: 4.7.6 Chunk 4

dim(Caravan)

with(Caravan, summary(Purchase))

348 / 5822

```

```{r}
#| label: 4.7.6 Chunk 5

standardized.X <- scale(Caravan[, -86])

var(Caravan[, 1])

var(Caravan[, 2])

var(standardized.X[, 1])

var(standardized.X[, 2])

```

```{r}
#| label: 4.7.6 Chunk 6

test <- 1:1000

train.X <- standardized.X[-test, ]
test.X <- standardized.X[test, ]

train.Y <- Caravan$Purchase[-test]
test.Y <- Caravan$Purchase[test]

set.seed(1)

knn.pred <- knn(train.X, test.X, train.Y, k = 1)

mean(test.Y != knn.pred)

mean(test.Y != "No")

```

```{r}
#| label: 4.7.6 Chunk 7

table(knn.pred, test.Y)

9 / (68 + 9)

```

```{r}
#| label: 4.7.6 Chunk 8

knn.pred <- knn(train.X, test.X, train.Y, k = 3)
table(knn.pred, test.Y)
5 / 26

knn.pred <- knn(train.X, test.X, train.Y, k = 5)
table(knn.pred, test.Y)
4 / 15

```

```{r}
#| label: 4.7.6 Chunk 9

glm.fits <- glm(Purchase ~ .,
                data = Caravan,
                family = binomial,
                subset = -test)
glm.probs <- predict(glm.fits, Caravan[test, ], type = "response")

glm.pred <- rep("No", 1000)
glm.pred[glm.probs > .5] <- "Yes"
table(glm.pred, test.Y)

glm.pred <- rep("No", 1000)
glm.pred[glm.probs > .25] <- "Yes"
table(glm.pred, test.Y)
11 / (22 + 11)

```

## Problem 2: Recidivism Prediction

Just like in Lab 5, we will use additional classification models to predict whether an arrested person will reoffend within two years.

### Part a (Code: 3.5 pts; Explanation: 1 pt)

Choose an accuracy metric to optimize (e.g., AUC, mean log-loss). Using 5-fold cross-validation on the training set, determine and justify an "optimal" value for the number of nearest neighbors in a k-nearest neighbors model. Support your decision by creating a plot showing the average accuracy as a function of `k`.

If you are using "regular R", Euclidean distance is used by default, and you will need two `for` loops (an outer loop over the folds and an inner loop over the possible values of `k`). To avoid confusion with `k` in k-fold vs `k` in k-nn, use `v` as the fold index.

If you are using `tidymodels`, you should create a `model` specifying the algorithm and the parameter to tune, then create a grid of possible parameter values and tune over it. You should set `dist_power = 2` and not worry about tuning it.

#### Solution

```{r}
#| label: Monster Chunk

# Setting up model for knn classification
knn_model <- nearest_neighbor(
  mode = "classification",
  neighbors = tune(),
  dist_power = 2
)

# Standard recipe with normalization
knn_recipe <- recipe(
  recid ~ age_at_screening + sex + priors,
  data = compas_train
) |>
  step_dummy(all_nominal_predictors()) |>
  step_normalize(all_numeric_predictors())

# Basic Workflow
knn_wflow <- workflow() |>
  add_model(knn_model) |>
  add_recipe(knn_recipe)

set.seed(676767)

# Defines/plans 5 Fold Cross Validation, strata is response variable
knn_kfold <- vfold_cv(compas_train,
                      v = 5,
                     strata = recid)

# Parameter grid, sets up which k's to try
# Take one value, 5-fold CV, compute metric, then repeat
knn_grid <- expand.grid(
  neighbors = seq(2, 1502, by = 50)     # Keep testing until you get the roc auc pattern we are familiar
)

# Performs 5 Fold and computes metric, averages the results across folds
knn_tune <- tune_grid(
  knn_wflow,     # workflow to tune
  resamples = knn_kfold,     # tibble with instructions to k-fold cv
  grid = knn_grid,     # grid of hyperparameters values to search over
  metrics = metric_set(roc_auc)
)

# Plot
autoplot(knn_tune)

# Best k
best_k <- select_best(knn_tune, metric = "roc_auc")
best_k

```

I used ROC AUC for the accuracy metric and with 5-fold cross-validation. Based on the plot and code, its peak is around 152 nearest neighbors, then it starts to level off. That is why I chose K to be 152, though in the code I left it as `best_k <- select_best(knn_tune, metric = "roc_auc")` for ease of use and stuff.

### Part b (Code: 1.5 pts)

Fit a k-nearest neighbors model on the full `compas_train` training set using the value of `k` selected in Part a.

Then, using this model, predict whether each arrested person in the `compas_test` dataset will be arrested within two years of release (`recid == yes`). Also, create a `compas_knn_predictions` data frame or tibble containing both the predicted and actual classes and the predicted probabilities of "yes" and "no". 

Remember to re-scale the predictor variables in your test set based on the full training set!

#### Solution

```{r}
#| label: Fitting

knn_model_final <- nearest_neighbor(
  mode = "classification",
  neighbors = best_k$neighbors,     # hard coded instead of select best is done here
  dist_power = 2
)

knn_wflow_final <- workflow() |>
  add_model(knn_model_final) |>
  add_recipe(knn_recipe)

knn_fit_tuned <- fit(knn_wflow_final,
                     data = compas_train)

compas_knn_predictions <- knn_fit_tuned |>
  augment(new_data = compas_test)

compas_knn_predictions

conf_mat(
  compas_knn_predictions,
  truth = recid,
  estimate = .pred_class
)

```


### Part c (Code: 1.5 pts)

Fit a Naive Bayes model on the `compas_train` training set predicting `recid` from `age_at_screening`, `priors`, and `sex`. Remember that you will have to convert `sex` to a factor variable (or use `step_dummy`) if you want to use tidymodels!

Then, using this model, predict whether each arrested person in the `compas_test` dataset will be arrested within two years of release (`recid == yes`). Also, create a `compas_nb_predictions` data frame or tibble containing both the predicted and actual classes and the predicted probabilities of "yes" and "no". 

#### Solution

```{r}
#| label: Naive Bayes Model + Fit

nb_model <- naive_Bayes(
  mode = "classification",
  engine = "klaR"
  # both are defaults
)

nb_recipe <- recipe(
  recid ~ age_at_screening + priors + sex,
  data = compas_train
)
# Requires factor variables of categorical - NO DUMMY VARIABLES!

nb_wflow <- workflow() |>
  add_model(nb_model) |>
  add_recipe(nb_recipe)

nb_fit <- nb_wflow |>
  fit(data = compas_train)

compas_nb_predictions <- nb_fit |>
  augment(new_data = compas_test)

compas_nb_predictions

conf_mat(
  compas_nb_predictions,
  truth = recid,
  estimate = .pred_class
)

```


### Part d (Code: 1.5 pts)

Fit a linear discriminant analysis model on the training set predicting `recid` from `age_at_screening`, `priors`, and `sex`. 
Then, using this model, predict whether each arrested person in the `compas_test` dataset will be arrested within two years of release (`recid == yes`). Also, create a `compas_lda_predictions` data frame or tibble containing both the predicted and actual classes and the predicted probabilities of "yes" and "no". 

#### Solution

```{r}
#| label: LDA, Model + Fit + Predict

lda_model <- discrim_linear(
  mode = "classification",      # default
  engine = "MASS"     # default
)

lda_wflow <- workflow() |>
  add_model(lda_model) |>
  add_recipe(nb_recipe)

lda_fit <- lda_wflow |>
  fit(data = compas_train)

compas_lda_predictions <- lda_fit |>
  augment(new_data = compas_test)

compas_lda_predictions

conf_mat(
  compas_lda_predictions,
  truth = recid,
  estimate = .pred_class
)

```

### Part e (Code: 1.5 pts)

Fit a quadratic discriminant analysis model on the training set predicting `recid` from `age_at_screening`, `priors`, and `sex`. Remember that you will have to convert `sex` to a factor variable (or use `step_dummy`) if you use tidymodels!

Then, using this model, predict whether each arrested person in the `compas_test` dataset will be arrested within two years of release (`recid == yes`). Also, create a `compas_qda_predictions` data frame or tibble containing both the predicted and actual classes and the predicted probabilities of "yes" and "no". 

#### Solution

```{r}
#| label: QDA, Model + Fit + Predict

qda_model <- discrim_quad(
  mode = "classification",      # default
  engine = "MASS"     # default
)

qda_wflow <- workflow() |>
  add_model(qda_model) |>
  add_recipe(nb_recipe)

qda_fit <- qda_wflow |>
  fit(data = compas_train)

compas_qda_predictions <- qda_fit |>
  augment(new_data = compas_test)

compas_qda_predictions

conf_mat(
  compas_qda_predictions,
  truth = recid,
  estimate = .pred_class
)

```

## Problem 3: Evaluating Probabilistic Predictions

### Part a (Code: 0.5 pts; Explanation: 1 pt)

Using the `brier_class` function, obtain the Brier score for each model.

Which of the four models performs best by this measure? Which performs worst? Explain.

#### Solution

```{r}
#| label: Brier Class

brier_class(compas_knn_predictions, truth = recid, .pred_yes)
brier_class(compas_nb_predictions, truth = recid, .pred_yes)
brier_class(compas_lda_predictions, truth = recid, .pred_yes)
brier_class(compas_qda_predictions, truth = recid, .pred_yes)

```

For Brier Class, the lowest brier score is the best while the highest is the worst. So, our best model is LDA with a brier score of about .2038. Our worst model is QDA with a brier score of about .2097.

### Part b (Code: 0.5 pts; Explanation: 1 pt)

Using the `mn_log_loss` function, obtain the cross-entropy/log loss for each model.

Which of the four models performs best by this measure? Which performs worst? Explain.

#### Solution

```{r}
#| label: mn_log_loss

mn_log_loss(compas_knn_predictions, truth = recid, .pred_yes, event_level = "first")
mn_log_loss(compas_nb_predictions, truth = recid, .pred_yes, event_level = "first")
mn_log_loss(compas_lda_predictions, truth = recid, .pred_yes, event_level = "first")
mn_log_loss(compas_qda_predictions, truth = recid, .pred_yes, event_level = "first")

```

For mn_log_loss, the lowest .estimate value is the best while the highest is the worst. So, our best model is LDA with an estimate value of about .6033. Our worst model is QDA with an estimate value of about .6782.
