---
title: Final Project
---

Remember to replace all of the instructions with your text!

## Summary

This section should include:

* Your main question or objective
My main objective is to see if it is possible to somewhat be able to predict a Pokemon's type based off of their base stats.

* The main insight(s) from your project
x

* A figure that supplements that insight, along with accompanying text describing the figure. This figure can be a cleaned-up version of a visualization you produced in EDA or a visualization using your final model.
x

## Motivation and Context

```{r}
#| label: do this first
#| echo: false
#| message: false

# change this to the location of your Quarto file containing your project; then delete this comment
here::i_am("final_project.qmd")
```

My investigation revolves around the Pokemon franchise, more specifically the statistical data of certain Pokemon in their games. This statistical data consists of their health, attack, defense, special attack, special defense, and speed. This info determines the raw strength of a Pokemon along with their level they may be. Each Pokemon has a type, like fire and water and such. With a total of 18 types, there are many differences between each Pokemon. The purpose of types is that it creates strategic advantages as some are stronger than others, some may be immune to others, and so on and so forth. Some Pokemon may even have two types, but I am focusing mostly on their primary type.

My initial thoughts were to determine if certain types tend to have a superior average total of stats. That was kind of simple so I delved into the idea that it may be possible that types might have patterns in their base stats, like rock types might have better defense. This eventually grew to our final idea, can we predict a Pokemon's type based off stats?

So, I am investigating if we can predict a Pokemon's primary type based off their base stats. This will be done by  Multinomial Logistic Regression. Accurate predictions may very well imply that certain patterns exist between primary typing and base stats.

I am investigating out of pure curiosity and passion for Pokemon and to find out if I can truly apply what i have learned in class towards independent "research" and datasets. I do also find it interesting if we can get a somewhat accurate type predictor based of off base stats.

## Packages Used In This Analysis

This section lists the packages used in the analysis. You should have a single chunk that loads the packages followed by a description of why you are using each package. Please use individual package names rather than tidyverse and tidymodels, to make it clear which specific packages are being used.

```{r}
#| label: load packages
#| message: false
#| warning: false

library(here)        # File path management and here::i_am()
library(readr)       # Reading data files such as CSVs
library(dplyr)       # Data wrangling: select, filter, mutate, summarise, arrange
library(rsample)     # Train/test splits and cross-validation
library(ggplot2)     # Data visualization and graphs
library(tidyr)       # Data reshaping: pivot_longer()
library(broom)       # augment() and tidying model outputs
library(factoextra)  # Cluster visualization: fviz_cluster()
library(kknn)        # K-nearest neighbors engine
library(forcats)     # Factor reordering functions like fct_infreq() and fct_reorder()
library(nnet)        # Multinomial logistic regression engine
library(recipes)     # Preprocessing and recipe steps like step_normalize()
library(workflows)   # Combining models and recipes into workflows
library(parsnip)     # Model specifications: nearest_neighbor(), multinom_reg()
library(tune)        # Hyperparameter tuning: tune_grid(), select_best()
library(yardstick)   # Model evaluation metrics: mn_log_loss()
library(pokemon)     # Pokemon dataset

```

| Package | Use |
|-------------------------------|----------------------------------------|
| [here](https://github.com/jennybc/here_here) | to easily load and save data |
| [readr](https://readr.tidyverse.org/) | to import the CSV file data |
| [dplyr](https://dplyr.tidyverse.org/) | to massage and summarize data |
| [rsample](https://rsample.tidymodels.org/) | to split data into training and test sets and for cross-validation |
| [ggplot2](https://ggplot2.tidyverse.org/) | to create nice-looking and informative graphs |
| [broom](https://broom.tidymodels.org/) | to augment data |
| [factoextra](https://cran.r-project.org/web/packages/factoextra/index.html) | to visualize clusters |
| [kknn](https://cran.r-project.org/web/packages/kknn/index.html) | to use its engine for K-Nearest Neighbors |
| [forcats](https://forcats.tidyverse.org/) | to factor reorder |
| [nnet](https://cran.r-project.org/web/packages/nnet/index.html) | to use its engine for Multinomial Linear Regression |
| [recipes](https://recipes.tidymodels.org/index.html) | to build recipes for workflows |
| [workflows](https://workflows.tidymodels.org/index.html) | to combine models and recipes to build workflows |
| [parsnip](https://parsnip.tidymodels.org/index.html) | to compute Multinomial Linear Regression |
| [tune](https://tune.tidymodels.org/index.html) | to select best tunes |
| [yardstick](https://yardstick.tidymodels.org/index.html) | to use mn_log_loss() for model evaluation |
| [pokemon](https://cran.r-project.org/web/packages/pokemon/readme/README.html) | to use its data set |

## Data Description

  An observation in this dataset presents a Pokemon as featured in their debut game, rather than any later appearances or alternate form (Deoxys). The various game stats data included are their base stats (HP, attack, defense, special attack, special defense, and speed) as featured in their debut game, along with a total base stat value, height, weight, Pokedex ID, and their generation id. There are 649 observations with 15 different variables.

  This CRAN dataset was collected by William Amorim and published in 2023. Based on linked sources, I would assume stats data was collected from Bulbapedia, a reliable and community-driven online encyclopedia dedicated to the Pokemon franchise, and from a GitHub that hosts data for Pokemon as well. A separate GitHub was used to link to Pokemon images.

  Indexing issues are caused by Mega Pokemon and by alternate versions of Pokemon, so I subsetted the dataset. Another reason for this was due to my lack of interest in the Pokemon past the 5th generation. So, Pokemon removed were any beyond the 5th generation. A few variables were removed too, such as 2 columns with urls that link to images, 3 columns that contained hex codes to their main colors, species id variable, and the 2 egg group variables. This is due to the fact that they will be useless here. The species id variable was removed since it was the same has the id variable, and egg groups are just irrelevant to my narrative. I also added a total stats variable that sums up all 6 of their base stats for a singular variable.

## Data Wrangling

```{r}
#| label: Data Wrangling 

pokemon <- pokemon |>
  dplyr::select(
    id, pokemon, height, weight, base_experience, type_1, type_2,
    hp, attack, defense, special_attack, special_defense, speed,
    generation_id
  ) |>
  dplyr::filter(generation_id < 6) |>
  mutate(total_stats = hp + attack + defense + special_attack + special_defense + speed)

set.seed(67)

pokemon_split <- initial_split(pokemon, prop = .8)

pokemon_train <- training(pokemon_split)
pokemon_test <- testing(pokemon_split)

```

## Exploratory Data Analysis

This section should include the code, figures, and accompanying text for the exploratory data analysis that is relevant to your objective. You should be able to just clean up the EDA section from Project 2.

## Modeling

The section should include the code, output, and accompanying descriptions for *one* supervised or unsupervised learning model, as appropriate to your question or objective. The descriptions should include some conceptual or mathematical background explaining how the model works.

You should be able to just clean up one of the modeling sections from Project 2.

```{r}
#| label: Multinomial Logistic Regression Setup

set.seed(67)

# Convert response to factor
pokemon_train$type_1 <- as.factor(pokemon_train$type_1)

multi_model <- multinom_reg(
  mode = "classification",
  engine = "nnet"
)

multi_recipe <- recipe(
  type_1 ~ hp + attack + defense + special_attack + special_defense + speed,
  data = pokemon_train
)


multi_wflow <- workflow() |>
  add_model(multi_model) |>
  add_recipe(multi_recipe)

multi_fit <- fit(multi_wflow, data = pokemon_train)

multi_fit

```

## Insights/Conclusion

Pick either "Insights" or "Conclusion" as the section title depending on what makes more sense.

Describe the primary findings from your EDA and modeling sections as appropriate to the main question/objective. You can reproduce the figure from your summary if that helps.

### Limitations and Future Work

This section should include further exploration of the limitations of your modeling approach. Some avenues of exploration include:

-   Who/what is being poorly predicted by your model? Why?
-   How might someone improve the data collection or analysis in the future?
-   Are the assumptions behind your analytical methods justified?
-   Are there ethical concerns about how the data was collected or how the results of the analysis could be used in the real world?

## Extra Credit

If you went to any off-campus events this semester that involved statistics or math in some way (e.g., DataFest, GSB off-campus events, etc.) or attended on-campus events that brought in potential future colleagues/mentors from off-campus (e.g., grad school/career fairs, TSC events, etc.), you can write a paragraph at the end of the reflection telling me what you went to and what you got out of it (if it was a waste of time, tell me why).

For each paragraph, you get an extra 5 points on this assignment, up to 20 points.




